Extrusion-modulated DnaA activity oscillations coordinate DNA replication with biomass growth

Version 1

Abstract

Abstract

Robust control of DNA replication is fundamental to bacterial proliferation. In Escherichia coli, replication initiation is thought to be regulated by oscillations in DnaA activity, driven by DnaA-chromosome interactions that differ among leading models. However, direct evidence linking these oscillations to replication initiation has been lacking, and existing models fail to explain the observed decoupling of replication initiation from dnaA expression. Here, we establish a direct link between DnaA activity and replication initiation by demonstrating robust oscillations in DnaA activity, which peak precisely at replication initiation across diverse growth conditions and genetic perturbations. Notably, these oscillations persist even when dnaA transcription remains constant, suggesting a regulatory mechanism that modulates DnaA activity independently of its expression. Additionally, we propose an extrusion model in which DNA-binding proteins sense biomass-DNA imbalance and extrude DnaA from the chromosome to trigger replication, overcoming limitations of existing models. Consistent with this model, perturbation of the nucleoid-associated protein H-NS modulates DnaA activity and replication timing, supporting its mechanistic validity.

Introduction

Bacterial proliferation requires precise coordination between DNA replication and biomass growth to maintain cellular homeostasis. While biomass accumulates exponentially, DNA synthesis proceeds in a stepwise manner, creating temporal imbalances that demand continuous correction throughout the cell cycle (Fig. 1A-B). Failure to synchronize these processes risks catastrophic outcomes, such as growth defects, incomplete replication, unequal genetic segregation, and cell death (Boye et al., 1996; Khodursky et al., 2015; Lu et al., 1994; Mäkelä et al., 2024; Zaritsky & Pritchard, 1971). Central to this synchronization in Escherichia coli is the initiator protein DnaA, a DNA-binding protein that binds specifically to an asymmetric 9-bp DnaA-box, whose consensus sequence is TTWTNCACA (Hansen & Atlung, 2018; Speck et al., 1999). Although DnaA binding to the replication origin (oriC) triggers replication initiation (Dong et al., 2023; Kaguni, 2011; Ohbayashi et al., 2020; Ozaki & Katayama, 2012; Sekimizu et al., 1987), how DnaA activity is dynamically regulated to align replication initiation with cellular growth, particularly when DNA content and biomass diverge, remains a fundamental unresolved question in cell cycle biology.

Coordination of biomass growth and DNA replication through replication initiation and DnaA activity oscillations.

(A) Schematic representation of the discrepancy between biomass accumulation and DNA replication. While biomass grows exponentially, DNA synthesis progresses linearly, necessitating replication initiation events to maintain coordination. (B) Mechanistic model for biomass-DNA coordination in bacteria. A molecular sensor detects deviations between cell mass and DNA content, transmitting this information to regulatory controllers that compensate by either increasing DNA replication or restricting biomass accumulation. (C) Illustration of cyclic DnaA activity oscillations aligning with replication initiation to ensure precise cell cycle control.

Two competing models have dominated the field. The titration model posits that DnaA activity is governed by the availability of free DnaA, which accumulates until it surpasses the buffering capacity of chromosomal DnaA-boxes (Christensen et al., 1999; Hansen et al., 1991). Here, replication initiation resets the system by increasing DNA content (and thus DnaA-boxes), titrating excess DnaA. In contrast, the switch model emphasizes the ATP/ADP-bound states of DnaA: DnaA-ATP, synthesized during biomass growth, triggers initiation, while hydrolysis to DnaA-ADP following replication resets the cycle(Berger & Wolde, 2022; Donachie & Blakely, 2003; Fu et al., 2023; Hansen & Atlung, 2018; Katayama et al., 2017). Crucially, the titration model depends on ongoing DnaA synthesis to reflect biomass accumulation, whereas the switch model relies on nucleotide-state transitions, independent of total DnaA levels. Despite these mechanistic differences, both models consider that replication initiation is regulated by oscillations in DnaA activity, driven by DnaA-chromosome interactions (Fig. 1C). However, direct evidence linking these oscillations to replication initiation remains lacking. Moreover, neither model fully explains recent experimental observations: replication initiation persists for multiple generations even when dnaA transcription is completely inhibited (Knöppel et al., 2023). This discrepancy suggests an uncharacterized mechanism that senses biomass-DNA imbalance and directly modulates DnaA activity.

Here, using a synthetic reporter coupled with single-cell mRNA fluorescence in situ hybridization (FISH), we demonstrate that DnaA activity peaks precisely at initiation across diverse growth conditions and genetic backgrounds, even when dnaA transcription is held constant. DnaA activity oscillations persist independently of transcriptional feedback, pointing to a post-translational regulatory mechanism. We propose an extrusion model in which nucleoid-associated proteins (NAPs) sense biomass-DNA imbalance and dynamically extrude DnaA from the chromosome, liberating it to activate oriC. This model not only explains the decoupling of replication timing from dnaA expression but also integrates disparate experimental findings. Perturbations of the NAP, H-NS for instance, modulate DnaA activity and replication timing, directly supporting the model’s predictions. By unifying regulatory principles across transcription, protein activity, and chromosome dynamics, our findings redefine the paradigm for bacterial replication control.

Results

Development of a DnaA activity reporter system

It has been known that changing DnaA expression levels affects DnaA activity, and further changes initiation mass (Berger & Wolde, 2022; Løbner-Olesen et al., 1989; Zheng et al., 2020). To investigate this further, we engineered a dnaA-titratable strain, where dnaA transcription is controlled by an inducible Ptet promoter (Lutz & Bujard, 1997) (Fig. 2A). This system enabled precise modulation of dnaA mRNA levels (0.25 to 6 times the wild-type level) via anhydrotetracycline (aTc) induction (0.75– 20 ng ml−1) without altering growth rate or cell size (Fig. 2B–D). Strikingly, increasing DnaA expression doubled population-averaged oriC copy numbers and reduced initiation mass (the cell volume per oriC at replication initiation) by 50 % (Fig. 2E-F). The observed relationship between DnaA expression and initiation mass deviated significantly from both the titration and switch models (Fig. 2G), suggesting unaccounted regulatory mechanisms.

Construction and characterization of the dnaA-titratable strain.

(A) Schematic of the dnaA-titratable strain. A dnaA gene under the control of the Ptet promoter was inserted near oriC and regulated by a Ptet-tetR feedback loop integrated at the intS locus, enabling fine-tuned expression control. The native dnaA gene was replaced with a kanamycin resistance cassette (kanr). (B–F) Characterization of dnaA-titratable cells (red open circles) grown in rich defined medium with glycerol (M6) under varying aTc concentrations. Measured parameters include dnaA mRNA levels (B), growth rate (C), population-averaged cellular mass (D), population-averaged oriC numbers (E), and initiation mass (F). Wild-type MG1655 values (black triangles) are shown for comparison. dnaA mRNA levels, normalized to wild-type MG1655, increased with aTc concentration from 0.75 to 20 ng·ml−1. Cellular mass was determined by OD600 divided by cell number concentration. oriC copy numbers were measured using a run-out assay, and initiation mass was calculated as the ratio of cellular mass to oriC numbers. Data represent means ± SD (n = 5 biological replicates). (G) Relationship between relative initiation mass and relative dnaA mRNA levels, compared with predictions from the initiation titration model (blue line) and the switch model (purple line). Experimental data are overlaid for validation.

To dissect DnaA activity dynamics, we constructed a library of synthetic promoters (n = 67) by replacing tetO operators in Ptet with DnaA-boxes (Fig. S1). To assess the degree of DnaA activity-mediated repression, we measured the fold change in GFP expression driven by each synthetic promoter under low and high levels of dnaA induction (Fig. 3A). Six synthetic promoters exhibited over 8-fold DnaA-mediated repression, whereas the promoter lacking DnaA-boxes (Pcon) showed none, and the well-characterized native dnaA promoter (Pnative)(Braun et al., 1985; Saggioro et al., 2013; Speck et al., 1999) displayed approximately two-fold repression (Fig. 3B; Fig. S1). Psyn66, which contains three strong and one weak DnaA-box, exhibited >8-fold repression under high DnaA induction (Fig. 3B). Psyn66’s design enables responsiveness to both free DnaA levels (via strong boxes) and the ATP/ADP ratio (via differential binding to the weak box)(Grimwade & Leonard, 2021; Katayama et al., 2017; Speck et al., 1999). Dose-response assays confirmed that Psyn66 was progressively repressed across a 40-fold range of dnaA expression (Fig. 3C), with no interference from SeqA-mediated sequestration (Fig. 3D-E). These results validated Psyn66 as a robust and specific reporter of DnaA activity.

Development of a DnaA activity reporter system.

(A) Schematic of promoter design and screening. Sixty-seven synthetic promoters were constructed by inserting various DnaA-boxes around the promoter core to drive gfp expression in a dnaA-titratable strain, where dnaA expression was regulated by aTc concentration. After pre-cultivation, GFP fluorescence per OD600 was measured using a microplate reader in cells grown under low (0.5 ng·ml−1) or high (50 ng·ml−1) aTc concentrations to determine repression fold-change. (B) Repression fold-change of synthetic promoters. Pcon (a promoter lacking DnaA-boxes) and Pnative (the endogenous dnaA promoter) served as negative and positive controls, shown in gray and yellow, respectively. (C) Response curves of three promoters to varying dnaA expression levels, with their promoter architectures shown on the right. Promoter activity was assessed by relative gfp mRNA levels, normalized to the lowest dnaA expression condition. Schematic (D) and response curves (E) of Psyn66 and Pnative responses to SeqA in the seqA-titratable strain. Promoter activity was quantified from gfp transcript levels in seqA-titratable cells containing the Psyn66-GFP plasmid, normalized to the lowest seqA expression level. All cells were grown in rich defined medium supplemented with glycerol across different aTc concentrations. Data represent mean ± SD from 3 biological replicates (B,C,E).

Decoupling DnaA activity oscillations from dnaA transcription fluctuations

To investigate DnaA activity dynamics throughout the cell cycle, we linked Psyn66 activity to cell size, a surrogate for cell-cycle progression, using single-cell mRNA FISH(Skinner et al., 2013) to quantify lacZ mRNA levels in MG1655Δlac cells harboring the Psyn66-lacZ plasmid (Fig. 4A). A control strain harboring a promoter-less lacZ construct (Pneg) showed undetectable fluorescence signals (Fig. 4A), confirming the specificity of lacZ mRNA detection. Quantitative analysis revealed that Psyn66-driven mRNA levels fluctuated approximately 3-fold over the cell cycle, whereas the DnaA-unresponsive constitutive promoter (Pcon) exhibited stable expression (Fig. 4B). However, fluctuations in mRNA concentration may not directly correspond to changes in DnaA activity due to global factors such as plasmid copy number and RNA polymerase concentration (Balakrishnan et al., 2022; Bintu et al., 2005a; Bintu et al., 2005b). To account for these variables, we used a Pcon-lacZ plasmid to estimate global effects on mRNA concentration and calculated DnaA activity as inversely proportional to . In wild-type cells (MG1655Δlac) harboring the synthetic reporter system (Fig. 4C), fluctuations in DnaA activity, as denoted by , displayed approximately a 3-fold variation across the cell cycle (Fig. 4D), underscoring the substantial cell cycle-dependent oscillations in DnaA activity.

DnaA activity oscillations decoupled from dnaA transcription fluctuations.

(A) Representative lacZ mRNA FISH images of MG1655 Δlac cells transformed with lacZ expression plasmids driven by Psyn66 (DnaA-boxes around promoter core), Pcon (no DnaA-box around promoter core), or Pneg (mutated promoter core). Yellow outlines indicate cell boundaries identified from phase-contrast images. (B) Relative lacZ mRNA concentrations driven by Psyn66 (left) and Pcon (right) across different cell volumes. Relative concentrations were determined from volume-specific lacZ mRNA fluorescence intensities, normalized to the population average. Volume-binned data for Psyn66 ([mZ] (Psyn66)) and Pcon ([mZ] (Pcon)) are shown as open circles and were used to calculate ksyn66. (C) Schematic of a strain with autoregulated dnaA transcription carrying a DnaA activity reporter plasmid. Cell cycle-dependent fluctuations in relative DnaA activity (D) and relative dnaA mRNA concentrations (E) in cells from panel C, grown in rich defined medium supplemented with glucose. Relative DnaA activity , calculated from volume-binned data in panel B, was smoothed and plotted as a red curve (D). Relative dnaA mRNA concentrations were determined from volume-specific dnaA mRNA fluorescence intensities, normalized to the population average, with volume-binned data shown as open circles (E). (F–H) Same as panels C–E, but for cells with aTc induced dnaA transcription, grown in rich defined medium supplemented with glycerol under 2 ng·ml−1 aTc induction. More than 8,000 cells were analyzed per growth condition, and all error bars correspond to standard error of the mean (SEM).

We then examined whether the observed fluctuations in DnaA activity were linked to oscillations in dnaA transcription. Comparing the dynamics of dnaA mRNA and DnaA activity, we found that the peak in DnaA activity coincided with the trough in dnaA mRNA levels (Fig. 4D-E). To test whether dnaA transcription drives these oscillations, we eliminated transcriptional fluctuations using the dnaA-titratable strain. Despite constant dnaA mRNA levels (2 ng ml−1 aTc induction), DnaA activity was retained robust oscillations (Fig. 4F–H), demonstrating the decoupling of DnaA activity oscillations from transcriptional fluctuations. These suggest that post-translational regulation, not transcriptional feedback, governs the DnaA activity.

DnaA activity oscillations are tightly coupled to replication timing

To explore the relationship between DnaA activity oscillations and DNA replication initiation, we examined cells grown under various nutrient conditions supporting different doubling times (30–66 min). Significant fluctuations in DnaA activity were observed across all conditions (Fig. 5A). We also estimated the cell volume at replication initiation (Vi) based on population-averaged cell volume and oriC number (Bremer et al., 1979; Si et al., 2017; Zheng et al., 2016) (Fig. S2 A–C), marking this on the DnaA activity profiles. In a representative birth-to-division cell cycle (Pountain et al., 2024) (Fig. S2D), DnaA activity peaks consistently coincided with Vi, indicating a close correlation between DnaA activity and replication initiation (Fig. 5A).

Tight correlation between DnaA activity oscillations and DNA replication initiation.

(A) Cell cycle-dependent DnaA activity oscillations in wild-type cells across various growth conditions. DnaA activity is represented by . Volume-binned DnaA activity (red circles) was smoothed and plotted as a red curve. The representative birth-to-division cell cycle, defined as the cell volume doubling interval containing the majority of cells, is shaded in gray. The vertical line indicates the cell volume at replication initiation (Vi). (B) Cell cycle-dependent DnaA activity oscillations in dnaA-titratable cells grown in M6 medium with varying aTc concentrations. More than 8,000 cells were analyzed per growth condition, and error bars show the mean ± SEM. (C) Correlation between V∗ and Vi in wild-type (squares) and dnaA-titratable (circles) cells. V∗ represents the cell volume at the peak of DnaA activity within the representative birth-to-division cell cycle. The black line indicates equivalence between V∗ and Vi.

Further genetic perturbations, such as titrating DnaA expression, shifted DnaA activity fluctuations, yet the relationship between V∗ (the cell volume at maximal DnaA activity in a representative birth-to-division cell cycle) and Vi remained locked (Fig. 5B). To quantify this relationship, we compared Vi with V∗, revealing a strictly proportional relationship through the origin (slope= 1.0, R2 = 0.98), indicating equivalence between V∗ and Vi (Fig. 5C). These results confirm that DnaA activity oscillations are tightly correlated with replication initiation, independent of transcriptional or nutrient perturbations.

An extrusion model reconciles paradoxical initiation events

Very recently, Knöppel et al. reported that multiple rounds of initiation occurred even after DnaA synthesis is shut down (Knöppel et al., 2023). In their experiment, DnaA was supplied in excess before expression ceased (Knöppel et al., 2023), potentially allowing continued replication initiation due to residual DnaA. To address this issue, we shut down DnaA expression from its native level using the deactivated miniature Un1Cas12f1 (dUn1Cas12f1) system, guided by sgRNA to repress dnaA transcription (Fig. 6A). This system effectively diminished dnaA transcription to an extremely low level, yet total oriC numbers increased fourfold within 90 min, consistent with two rounds of replication initiation (Fig. 6B). These findings demonstrate that halting DnaA synthesis does not immediately abolish replication initiation.

An extrusion model explains DnaA shut-down dynamics.

(A) Genetic circuit of the deactivated CRISPR-Cas system for dnaA transcription shut-down. dnaA gene is targeted by a constitutively expressed sgRNA, while dUn1Cas12f1 expression is inhibited by TetR repressor. These transcription units are separated by terminators. The cassette was integrated into the chromosome near the oriC locus. DnaA shutdown is induced by the addition of aTc. (B) Time-course of relative dnaA mRNA levels (red line, left axis) and total oriC number (green line, right axis) following the addition of 50 ng·ml−1 aTc at time 0 (dashed line). dnaA mRNA levels were normalized to wild-type levels, and oriC numbers were normalized to their initial values. Error bars indicate mean ± SD (n = 3 biologically independent experiments). (C) Predicted increases in total oriC number during dnaA transcription shut-down based on three models: the titration model, switch model, and extrusion model. Shutdown was simulated by setting dnaA transcription to zero at time 0 (dashed line). (D) Schematic of the extrusion model. The model introduces extruder(s) as additional regulators of biomass-DNA coordination, complementing the role of DnaA (left). Increased binding of the extruder to DNA promotes the release of DnaA from DnaA-boxes (right). (E) Comparison of the relationship between relative initiation mass and relative dnaA mRNA levels from experimental data (Fig. 2F) and predictions of the extrusion model.

Both the titration and switch models predict a close relationship between DnaA activity oscillations and replication initiation(Berger & Wolde, 2022). However, they fail to account for the multiple rounds of initiation observed after DnaA synthesis was shut down. According to the titration model, initiation should cease immediately once DnaA expression is shut down, as the reduction in DnaA concentration would prevent it from reaching the threshold required to trigger new rounds of replication. In contrast, the switch model predicts infinite initiations following the cessation of DnaA expression (Fig. 6C). This is because the cessation of DnaA synthesis only reduce the production rate of DnaA-ATP and had no effect on the conversion rates between DnaA-ATP and DnaA-ADP. As a result, the ratio of DnaA-ATP to DnaA-ADP would temporarily decrease, causing a brief lag before initiation resumes indefinitely (Fig. 6C).

The failure of both models prompts us to revisit the regulatory mechanisms of replication initiation. In the titration model, a substantial fraction of DnaA proteins binds to DnaA-boxes along the chromosome, raising the possibility that these bound proteins serve as a reservoir of free DnaA, sustaining replication initiation even after DnaA synthesis ceases. We thus considered the possibility of some extruder(s) alongside DnaA to sense biomass-DNA deviation, and flush DnaA proteins away from the DnaA-boxes, thus releasing free DnaA to dynamically restore DnaA activity, even in the absence of ongoing DnaA synthesis (Fig. 6D).

In this proposed “extrusion model”, the extruder is constitutively expressed, synthesized at a constant rate throughout the cell cycle, accumulating in proportion to biomass growth, enabling replication initiation when free DnaA concentrations exceed the critical threshold, as described in the titration model(Hansen & Atlung, 2018; Hansen et al., 1991) (Fig. 6D). This model guarantees the long-term coordination of biomass growth and oriC abundance, as well as stable oscillations in free DnaA and oriC concentrations (Fig. S3 A). Notably, the extrusion model accounts for the persistence of replication initiation and free DnaA oscillations over multiple cell cycles after DnaA synthesis is halted (Fig. 6C; Fig. S3 B). Moreover, since the extruder mitigates the impact of perturbations on DnaA synthesis, the model predicts a quantitative relationship between initiation mass and DnaA expression levels that aligns with experimental observations (Fig. 6E). Furthermore, a stochastic implementation of the extrusion model shows the lack of correlation between successive replication initiation events (Fig. S3 C), as observed in recent studies (Si et al., 2019; Witz et al., 2019).

Perturbation of H-NS validates the extrusion mechanism

The extrusion model could be validated by changing the expression level of the extruder, which modulates DnaA activity by releasing DnaA from DnaA-boxes. Given their DNA binding properties, NAPs(Dillon & Dorman, 2010) are likely candidates for this extruder role. Therefore, we selected H-NS(Dorman, 2004), a major NAP, which is capable of promoting the release of DnaA from bound DnaA-boxes in vitro (Fig. S4), to experimentally assess the validity of the extrusion model. To this end, we constructed a hns-titratable strain, incorporating a reporting plasmid to capture changes in DnaA activity across various H-NS expression levels (Fig. 7A). Under steady-state cultivation, increasing aTc concentrations from 0 to 50 ng·ml-1 resulted in hns mRNA levels ranging from 0.25 to 5 times the wild-type levels, without affecting growth rate (Fig. S5 A-B). Within this range, DnaA activity increased by approximately 60 % (Fig. 7B), while the initiation mass decreased by 30 % (Fig. 7C; calculated with Fig. 5S C-D), in qualitative agreement with the predictions of the extrusion model (Fig. S5 E-F).

Titration of hns expression modulates DnaA activity and replication initiation.

(A) Genetic circuit of the hns-titratable strain. Expression of hns is controlled by a Ptet-tetR negative feedback loop integrated at the attB site, with the native hns coding sequence replaced by a kanamycin resistance gene. The plasmid contains Psyn66-mcherry and Pcon-gfp expression cassettes was used to assess DnaA activity. DnaA activity (B) and initiation mass (C) were characterized in M6 medium with varying hns expression levels during steady-state cultivation. (D) mRNA levels of hns (green circles) and dnaA (blue squares) relative to wild-type levels, along with DnaA activity (orange rhombus), were measured during hns shift-up. hns shift-up was induced by the addition of 50 ng·ml−1 aTc at time 0 (dashed line), following steady-state cultivation in M6 medium without aTc. (E) Dynamics of population-averaged oriC number (blue rhombus, left axis) and cellular mass (gray rhombus, right axis) during hns shift-up. Data represent mean ± SD from 3 biological replicates (B-E). (F) Cell cycle-dependent DnaA activity oscillations in hns-titratable cells cultivated in M1 and M6 media under varying aTc concentrations. Relative hns mRNA levels are indicated for each condition. More than 8,000 cells were analyzed for each condition and error bars show the mean ± SEM. (G) Correlation between the volume at maximal DnaA activity (V∗) and the volume at replication initiation (Vi) for hns-titratable cells grown in M1 (green down-pointing triangle) and M6 (red up-pointing triangle) media with varying hns expression levels. Data for dnaA-titratable cells (gray circles) and wild-type dnaA-autoregulated cells (Fig. 5C) are included for comparison.

Since H-NS is a global transcriptional regulator (Hommais et al., 2001), we wonder whether its effect on DnaA activity was mediated through dnaA transcription. To test this, we measured DnaA activity and dnaA mRNA levels during an H-NS shift-up. Upon addition of 50 ng·ml-1 aTc, hns levels increased more than 20-fold within a short period, followed by approximately 80 % increase in DnaA activity, with no significant change in dnaA mRNA levels (Fig. 7D). Additionally, cellular oriC content increased significantly within 30 minutes, while cell mass remained unchanged (Fig. 7E). These results suggest that the rapid increase in DnaA activity and accelerated replication initiation induced by H-NS upregulation is not driven by changes in dnaA transcription.

Leveraging the hns-titratable strain, we re-examined the tight correlation between DnaA activity oscillations and replication initiation. In two different growth media with varying hns levels, the peak of DnaA activity fluctuations (V∗) shifted in concert with the independently measured volume at replication initiation (Vi) (Fig. 7F). Plotting V∗ against Vi, revealed the same equivalency observed in wild-type and dnaA-titratable cells (Fig. 7G). Collectively, these results support the notion that DnaA activity oscillations, modulated by extrusion, coordinate DNA replication and biomass growth.

Discussion

In this study, we proposed chromosome-driven DnaA activity oscillations as a post-translational “rheostat” coordinating DNA replication with biomass growth in E. coli. By developing a SeqA-independent synthetic promoter reporter (Psyn66) on a plasmid, we circumvented confounding factors inherent to native systems (Pountain et al., 2024), thereby achieving the direct quantification of DnaA activity dynamics. Crucially, these oscillations persist when dnaA transcription is fixed, operate independently of SeqA-mediated repression, and peak precisely at replication initiation across all tested growth conditions. This reveals an essential paradigm: bacteria bypass transcriptional control to synchronize genome duplication with growth through spatial regulation of DnaA-chromosome interactions.

Very recently, Iuliani et al. also reported DnaA activity oscillations (Iuliani et al., 2024), which is a significant milestone, innovatively using a chromosomal promoter-reporter system and microfluidics for single-cell DnaA activity observations. They revealed key relationships between DnaA activity, cell size, and division. Our work, however, differs in multiple ways. Methodologically, we engineered a dnaA-titratable strain and synthetic promoters like Psyn66, enabling precise dnaA expression control and cleaner DnaA activity reporting. Our mRNA FISH approach uniquely compares DnaA activity and transcription at the single-cell level. By decoupling initiation volume measurement from DnaA monitoring, we resolve a 1:1 coupling between DnaA activity thresholds and replication initiation events. Regarding research questions, while Iuliani et al. focused on DnaA-cell size-division relationships, we addressed the decoupling of replication initiation from dnaA expression with our novel extrusion model.

Recently, an integrated “titration-switch” model has been proposed to improve replication cycle stability across different growth rates (Berger & Wolde, 2022; Fu et al., 2023). Although this integrated model predicts a relationship between initiation mass and dnaA mRNA levels that aligns closely with steady-state experimental data (Fig. S6 A), it fails to account for the additional replication initiation observed after dnaA shutdown (Fig. S6 B). By incorporating an extruder into this model, we successfully reproduced the additional replication initiation (Fig. S6 C), thus validating the extrusion concept and aligning with experimental observations. These findings highlight the critical role of extrusion in refining our mechanistic understanding of biomass-DNA coordination.

Mechanistically, we propose that NAPs sense biomass-DNA imbalances and extrude DnaA from chromosomal sites, converting physical displacement into biochemical activation, a spatial control strategy mirroring eukaryotic chromatin regulation. H-NS emerges as a prime extruder candidate given its DNA-condensing ability (Dame et al., 2000; Zimmerman, 2006) and replication-timing effects (Fig. 7). This functional multiplicity likely enhances robustness, allowing cells to maintain replication homeostasis despite fluctuating environments, an evolutionary advantage paralleling eukaryotic checkpoint system.

These findings redefine bacterial cell cycle control as a dynamic interplay between physical chromosome remodeling and metabolic sensing. By tethering DnaA activation to biomass accumulation through extrusion, cells achieve a critical feat: noise suppression via spatial filtering of transcriptional fluctuations. Such mechanistic parsimony, using chromosome structure to both store genetic information and regulate replication timing, may represent a fundamental principle of cellular organization. Indeed, parallels to eukaryotic systems are striking: DnaA extrusion resembles ORC licensing through chromatin accessibility (Fragkos et al., 2015), while biomass-DNA coupling evokes yeast size-control checkpoints (Xie et al., 2022).

Future studies could map the complete extruder interactome and elucidate how metabolic signals modulate their activity. Our synthetic reporter system provides a blueprint for dissecting the spatial regulation of DNA-binding proteins, with translational potential in antimicrobial development (targeting pathogen-specific extruders) and synthetic biology (engineering tunable replication circuits). More broadly, this work establishes physical chromosome remodeling as a central coordinator of cellular homeostasis, a concept bridging prokaryotic and eukaryotic cell cycle paradigms.

Materials and Methods

Plasmid and strain Construction

Strains used in this study were derived from E.coli MG1655. Information for all strains and plasmids are listed in Tables S1 and S2.

Plasmids for reporting DnaA activity

Plasmids expressing GFP, mCherry, or LacZ under synthetic promoters were constructed based on the pPT plasmid backbone (Zong et al., 2017). The lacZ amplified from the MG1655 genome and mcherry gene amplified from pMD19-hupA-mcherry plasmid, were assembled into the BspQI/BsrGI-digested pPT backbone via the One Step Cloning Kit (Vazyme, C112), to generate the pPT-lacZ and pPT-RFP plasmids, respectively. Synthetic promoters, generated by annealing the primer pairs listed in Table S3, were ligated directly into the BsaI-digested pPT, pPT-RFP or pPT-lacZ plasmid as required. The native dnaA promoter (Pnative) amplified from the MG1655 genome was digested with BsaI, and ligated into the BsaI-digested pPT plasmid. The Pcon-gfp cassette amplified from Pcon-GFP plasmid was assembled into the SpeI-digested Psyn66-RFP plasmid to generate the Psny66-Pcon-FPs plasmid. Primer information for amplifying DNA fragments is detailed in Table S4.

Plasmids for strain construction

For constructing the plasmid pMD19-Rhns, pMD19-RseqA and pMD19-RdnaA, the hns, seqA and dnaA gene were amplified from MG1655 genome, digested, and ligated into the same enzyme-digested pMD19-tetR plasmid. For constructing the plasmid p15A-RdnaA, the T3-Ptet-tetR-dnaA cassette, kanr gene and p15A-T1 fragment were amplified from plasmidpMD19-RdnaA, pEcCas and PZA31-Ptet-M2-GFP, respectively, and then assembled via MultiS One Step Cloning Kit (Vazyme, C113). The plasmid P_CRidnaA1 were obtained by ligating the annealed primer pairs to the BsaI-digested CPP00458 plasmid for the dnaA targeting. The chloramphenicol resistance gene Cmr amplified from the PZA31-Ptet-M2-GFP plasmid replaced the kanr gene in the pEcCas plasmid via λRed recombination method, resulting in the CmPcas plasmid. Primer information is detailed in Table S4.

Strain construction

The seqA-, and hns-titratable strains were constructed as previously described (Zheng et al., 2016). Briefly, after transforming pSIM5 plasmid into MG1655 cells, the Ptet-tetR-driven seqA or hns cassette was inserted at the attB site on the genome, and then the native gene was replaced with the kanr gene amplified from plkml plasmid.

For the construction of dnaA-titratable strain, we first inserted the Ptet-tetR cassette into the intS site via CRISPR/Cas9-λRed genome editing system (Jiang et al., 2015) by targeting 20-nucleotide sequence (tcttcctgcagaccagatcc). Then applying the λRed recombination systems in the CmPcas plasmid, the bla:Ptet-dnaA cassette amplified from the pMD19-RdnaA plasmid via overlap PCR, was inserted between the yidA and yidX genes, and the negative Pnative-dnaA was replaced with the T3-Kan-T1 cassette amplified from the p15A-RdnaA plasmid.

In order to delete the native lacZ gene in the genome of the RdnaA1, Rhns1 and MG1655 cells, the homologous recombination fragment amplified from the genome of CL1 strain, was used to knock out the native lac operon via CRISPR/Cas9-λRed editing system (Jiang et al., 2015) by targeting 20-nucleotide sequence (cttccggctcgtatgttgtg).

For the construction of DnaA shut-down strain (CRidnaA1), the PJ23119-sgRNAdnaA:PJ23100-tetR:Ptet-dUn1Cas12f1 cassette was amplified from the P_CRidnaA1 plasmid and flanked with homologous sequences to the asnA and viaA gene. This cassette was then inserted into the MG1655 genome using

CRISPR/Cas9-λRed editing system, targeting the 19-nucleotide sequence (aagccgcctgctcagacgc). The CPP00458 plasmid, kindly provided by the Xiongfei Fu lab, contains the Un1Cas12f1 variant, which carries mutations D143R, T147R, D326A, and K330R that abolish its DNA-cleaving activity. However, it retains the ability to bind DNA in a sgRNA-guided manner, thereby inhibiting the expression of the target gene.

All primers for the construction of DNA fragments are detailed in Table S4.

Growth conditions

Cells for strain construction were grown in in Luria-Bertani (LB) medium with appropriate antibiotics: ampicillin (100 μg·ml-1), kanamycin (50 μg·ml-1), spectinomycin (50 μg·ml-1), chloramphenicol (25 μg·ml-1), based on plasmid presents. For titratable strains, 10 ng·ml−1 anhydrotetracycline (aTc) was added during strain construction and seed culture preparation. Seed cultures were initiated by inoculating 3–6 individual colonies from an LB agar plate into 14 ml round-bottom test tubes containing 2 ml of the desired medium for overnight growth.

For physiological phenotypic characterization, all cells were cultivated at 37°C with shaking at 150 rpm in a water bath shaker. The dnaA-, seqA- and hns-titratable cells were characterized under varying aTc concentrations in rich defined medium supplemented with glycerol (M6). hns-titratable cells were also characterized in rich defined medium supplemented with glucose (M1). When strains contained reporter plasmid, 25 μg·ml−1 chloramphenicol was added. The autoregulated-dnaA strain (MGCL1) with a reporter plasmid, was tested in seven different media: M1 (MOPS + EZ + AUCG + Glucose), M2 (MOPS + EZ + Glucose), M3 (MOPS + EZ + AUCG + Gluconate), M6 (MOPS + EZ + AUCG + Glycerol),

M12 (MOPS + EZ + Glycerol), M13 (MOPS + CAA + Glycerol), and M18 (MOPS + Glucose). Detailed medium compositions were provided previously (Zheng et al., 2020). Steady-state growth was established as previously described(Zheng et al., 2020) by culturing cells for ∼10 generations in exponential growth phase with ∼15-fold dilution at OD600 ∼0.2, using the same medium supplemented with antibiotic or aTc as required. OD600 measurements were taken at consecutive time points using a spectrometer (Genesys 10s, Thermo Fisher Scientific), and the growth rate (or doubling time) was calculated by fitting the data to an exponential growth curve. Samples for determination of population-averaged cell mass and cellular origin number , quantification of mRNA levels, and mRNA fluorescence in situ hybridization (FISH) were collected when OD600 reached ∼0.2.

Measurements of population-averaged cellular mass , cellular oriC number , and initiation mass (mi)

The population-averaged cell mass , cellular oriC number , and initiation mass (mi) were calculated as described previously(Zheng et al., 2020). Briefly, was determined by dividing the OD600 by the cell concentration (OD600 ml per 109 cells). For cell counting, samples were collected in precooled cell count buffer (0.9% NaCl with 0.12% formaldehyde, filtered with a 0.22 μm filter) and stained with 1 μg·ml−1 DAPI in the same buffer. Cell number concentration was measured by counting the DAPI-stained particles using flow cytometry equipped with 405 nm laser (CytoFLEX, Beckman Coulter Life Sciences). The was calculated based on the classical run-out experiments(Zheng et al., 2020). In brief, cell suspensions were treated with 500 μg·ml−1 rifampicin and 30 μg·ml−1 cephalexin to inhibit new initiation and cell division, respectively, and maintained for 2-3 mass doublings to complete ongoing replication. Cells were fixed with 70% precooled absolute ethanol, and treated with stain buffer (20mM Tris-HCl, pH 8.0, 130 mM NaCl, 10 ng·ml−1 DAPI). Using the same flow cytometry, we measured the cellular origin distribution based on DAPI signaling, from which was calculated. mi was calculated based on the equation: (Bremer et al., 1979; Si et al., 2017; Zheng et al., 2016).

Screening of the synthetic promoter library

After transforming the reporter plasmid into dnaA-titratable cells, the seed culture was cultivated in M6 medium with 10 ng·ml-1 aTc, then diluted 1:180 in fresh medium containing either 0.5 or 50 ng·ml-1 aTc and cultivated at 37 °C and 800 r.p.m. in a digital thermo-shaker (AOSHENG) using flat-bottom 96-well plates (Corning). Following approximately 4 hours of pre-cultivation, the cultures were diluted 1:180 in fresh medium with the same aTc concentration. The OD600 and GFP fluorescence intensity were continuously monitored during cell growth using a Synergy H1 microplate reader (BioTek). The non-fluorescent strain (MG1655) cultured in 180ul of M6 medium for each experiment was used to subtract the background fluorescence. For OD600 between 0.1 and 0.2, the repression fold change was calculated by dividing the mean fluorescence intensity per OD600 of the cells grown with 0.5 ng·ml-1 aTc induction by the corresponding intensity for cells grown with 50 ng·ml-1 aTc induction.

Quantitative real-time PCR

Quantitative real-time PCR was performed using a protocol similar to previously described(Zheng et al., 2016). Briefly, RNA was stabilized using Bacteria Protect Reagent (Qiagen) and extracted using RNeasy mini kit (Qiagen). 500 ng RNA was reverse transcribed (TAKARA), and the resulting cDNA samples were diluted 1:25 with PCR-grade water. Quantitative real-time PCR was carried out on a qTower3 (Analytikjena) or CFX Connect system (Bio-Rad), using the following program: 30 s at 95 °C, followed by 40 cycles of denaturation (5 s at 95 °C), annealing, and elongation (30 s at 58 °C). Primer sequences were listed in Table S4. The rpoA gene was used as the reference gene.

lacZ and dnaA mRNA fluorescence in situ hybridization (FISH)

FISH for lacZ or dnaA mRNA was preformed using a method similar to previously described(Skinner et al., 2013). Briefly, antisense DNA oligo probes were designed against lacZ or dnaA, and synthesized with TAMRA labeling at the 3’ end. The sequences of probes targeting lacZ or dnaA are either as previously described(Skinner et al., 2013) or listed in Table S5. 27 ml steady-state growth culture (OD600∼0.2) was fixed by direct adding 3 ml of 37 % (vol/vol) formaldehyde, following by incubation for 30 min at room temperature with gentle mixing using a nutator. The fixed cells were collected by centrifugation at 600 g for 8 minutes, and washed twice with 1 ml RNase-free PBS. The cell pellet was resuspended in 300 μl DEPC water, and 700 μl precooled absolute ethanol was added to permeabilize the cells for 1 hour at room temperature in the nutator. After permeabilization, the cells were divided into two equal parts for characterization of lacZ or dnaA mRNA as needed. Hybridization was performed as described previously(Skinner et al., 2013), with the stringency of the hybridization and wash solutions adjusted by changing the concentration of formamide, 30 % for dnaA probes and 40 % for lacZ probes. After hybridization, the cells were washed with the appropriate wash solution and prepared for imaging.

Cell imaging and cell size parameters measurement

For imaging, 3 μl of cell suspension was pipetted onto the center of a 24×60-mm coverslip. A thin 1.5 % agarose pad (prepared with 1xPBS) was then placed on top of the cell suspension droplet. Fluorescent and phase images were acquired using an inverted microscope (IX-83, Olympus) equipped with a 100x oil objective (Olympus), an automated xy-stage (ASI, MS2000), and a sCMOS camera (Prime BSI, photometrics). Fluorescence excitation was performed as previously described (Xia et al., 2022). Solid-state lasers (Coherent OBIS: 405–100 LX, 561–100 LS) and Semrock emission bandpass filters (445/40 nm for DAPI and 598/25 nm for RFP) were used for excitation and collecting fluorescent signals.

Images were aligned with phase contrast channel by selecting at least six pairs of common control points and applying a geometric transformation using the MATLAB imwarp function. This allowed for adjustment of the cell masks in the RFP or DAPI channel. A customized ImageJ -based image-processing package, MicrobeJ (v5.13l1) plugin for Fiji (Ducret et al., 2016), was used to contour cells and calculate cell size parameters and fluorescence intensity, including the mean cell width, cell length, and cell area from phase-contrast images. The cell volume was calculated using a cylinder-plus-two-hemispheres geometry. The total RFP intensity, corrected for background subtraction per cell volume, was then used to estimate the concentration of the related mRNA level.

Characterization of cell cycle-dependent fluctuations in DnaA activity

Data filtering was first performed based on the intensity of the DAPI fluorescence signal. To eliminate non-cell regions resulting from incorrect segmentation of the phase contrast images, only DAPI-positive cells were included in the analysis. To further minimize the potential impact of variability in fixation, permeabilization, and wash efficiencies, which could affect the TAMRA fluorescence signal (representing the mRNA FISH signal), only cells with mean DAPI signals within the 95% confidence intervals were included in the subsequent analysis.

The concentration of the lacZ mRNA ([mZ]) can be expressed as [mZ] = J/δZ, where δZ is the degradation rate and J is the mRNA synthesis flux. J is in turn given by the promoter on-rate (k), the RNAP concentration ([RNAP]), and the concentration of the lacZ gene ([GZ]). Assuming that the factors δZ, [RNAP], and [GZ] are the same between the two strains with Psyn66 and Pcon driving lacZ, then the ratio of the mRNA concentration obtained at each point of the cell cycle gives the ratio of the promoter on-rates, kPsyn66 and kPcon. Since a constitutive promoter is non-responsive to DnaA activity, kcon is independent of cell cycle and we can obtain DnaA activity regulated promoter activity as , then the DnaA activity is denoted as .

The concentration of the lacZ mRNA were binned based on cell volume, with a bin size of 0.1 μm3, was used to determine [mZ] (Psyn66) and [mZ] (Pcon). The relation between [mZ] (Pcon) and cell volume was smoothed using the MATLAB smooth function, and from this, ksyn66 was deduced. Based on the DnaA activity oscillations, the cell volume where DnaA activity reaches maximum (V∗) was calculated during a representative birth-to-division cell cycle. The cell volume distribution was used to calculate the number of cells in each volume-doubling range. The range containing the highest cells was considered the representative birth-to-division cell cycle and was shaded in gray (Fig. S2 D). From this, V∗ was obtained.

Calculation of cell volume at the time of replication initiation (Vi)

Cell volume at the time of replication initiation (Vinit) was calculated based on population-averaged cell volume , and the population-averaged cellular oriC number . The equation used for this calculation is as follows: where vi is the initiation volume by definition, i.e. cell volume per oriC at the time of replication initiation(Bremer et al., 1979; Si et al., 2017; Zheng et al., 2016). was obtained by analyzing the phase contrast images, was obtained by run-out experiments(Zheng et al., 2020). . is the cellular oriC number at the time of replication initiation. denotes the floor function of log2 .

DnaA shut-down experiment

To shut-down DnaA expression, steady-state growth of the CRidnaA1 cells was first established after ∼10 generations of exponential growth in M1 medium. At OD600 between 0.15 and 0.2, 24 mL of cell culture was transferred to 12 mL of pre-warmed M1 medium containing 150 ng·mL−1 aTc to induce dUn1Cas12f1 expression at time 0. Under the guidance of sgRNAdnaA, the transcription of dnaA gene was inhibited. Samples were collected at various time points to determine population-averaged cellular mass , cellular oriC number , and to quantify mRNA levels. When the OD600 reached around 0.2, the cultures were diluted about 2-fold with pre-warmed M1 medium containing 50 ng·mL−1 aTc to maintain adequate nutrition. The steady-state growing MG1655 cells in M1 medium were also collected for quantification of wild-type dnaA mRNA levels. Three independent experiments were conducted.

H-NS shift-up experiment

The DnaA activity reporter plasmid Psyn66-Pcon-FPs was transformed into the hns-titratable strain (Rhns2), and cells were cultivated overnight in M6 medium containing 10 ng·mL−1 aTc and 25 μg·mL−1 chloramphenicol for seed culture. Seed cultures were washed and grown in the same medium without aTc for ∼10 generations of exponential growth. At an OD600 of ∼0.2, 40 mL of cell culture was transferred to 20 mL of pre-warmed same medium containing 150 ng·mL−1 aTc to induce H-NS expression at time 0. Samples were collected at time points -2, 1, 3, 5, 7, 10, 15, 20, 25, and 30 minutes for the determination of population-averaged cellular mass , cellular oriC number , and quantification of mRNA levels. Steady-state growing MG1655 in M6 medium were also collected for hns mRNA quantification. Three independent experiments were conducted.

Electrophoretic mobility shift assay (EMSA)

The genes encoding dnaA and hns from the MG1655 genome were individually amplified and cloned into the expression vector pET-28a using primers listed in Table S4. E. coli BL21 (DE3) carrying pET-28a-DnaA or pET-28a-H-NS was used for overproduction of His6-DnaA or His6-H-NS. For each protein purification, a 1 L culture was induced by the addition of 1 mM IPTG at OD600 = 0.4. After 1.5 hours of incubation, cells were harvested by centrifugation (5,000 rpm for 20 min at 4 °C). The pellet was resuspended in 50 ml of buffer A (50 mM Tris-HCl, pH 7.0, 500 mM NaCl, 0.5 mM PMSF, 5 % glycerol). Cell resuspension was lysed on ice for 30 minutes (3 seconds on, 7 seconds off) using a sonicator at 4 °C, and cell debris was removed by centrifugation (14,000 rpm for 60 min at 4 °C). Subsequently, the protein was purified from the soluble supernatant by Ni2+-affinity chromatography (HisTrap™ HP Columns; GE Healthcares) with a ÄKTA Pure Protein Purification System. Proteins were collected using gradient elution with buffer B [20 mM Tris-HCl, pH 7.5, 500 mM NaCl, 400 mM imidazole, 5 % glycerol]. Eluted proteins were further purified and concentrated by ultrafiltration using Amicon Ultra 10K (for H-NS) and 30K (for DnaA) filters (Millipore). Finally, after overnight dialysis (0.025 μm Membrane Filter; MF-Millipore) at 4 °C, the protein was stored in buffer C [45 mM Hepes-KOH, pH 7.6, 600 mM potassium glutamate, 1 mM DTT, 10 mM magnesium acetate, 0.5 mM EDTA, 20 % glycerol]. Judgement by SDS/PAGE with Coomassie Brilliant blue staining, the purity of each protein was >90 %. For the experiments where H-NS acts on the datA bound by DnaA, the purified DnaA, datA and ATP were first incubated in buffer D (20 mM Tris-HCl, pH 7.5, 150 mM potassium glutamate, 10 mM magnesium acetate, 8 mM DTT) at 20 °C for 10 min. Then the H-NS was added to buffer D (15 μl) which contained DnaA-datA complexes and incubated for additional 10 min at 20 °C. The final complexes were separated by electrophoresis in 2 % (wt/vol) agarose gel at 80 V for 70 min in Tris-Borate buffer, followed by staining with SYBR Safe (Invitrogen).

Titration model

In the titration model, we posit that the cell mass grows exponentially according to the differential equation , where V represents the volume of the cell and λ = 1.54 hr−1(corresponding to doubling time of 27 min) is the growth rate. Concurrently, DnaA protein synthesis occurs at a rate proportional to the cell’s mass, described by , with denoting the synthesis rate per unit volume. The chromosome’s structure is abstracted as a vector representing the copy numbers of uniformly distributed sites along its length. Specifically, DnaA-boxes, which are binding sites for DnaA proteins, are quantified by summing the copy numbers at their respective locations. In this model, 300 DnaA-boxes (Ab) are evenly distributed on the chromosome while an additional 100 DnaA-boxes are located near the origin of replication. This arrangement approximates a head-weighted distribution of DnaA-boxes and helps prevent over-initiation(Hansen et al., 1991). Binding of DnaA proteins to these DnaA-boxes allows us to calculate the concentration of free DnaA as [Af] = (A − Ab)/V, under condition of (A − Ab) ≥ 0. DNA replication commences once the concentration of free DnaA exceeds a critical threshold . 10 min sequestration time was set after replication initiation to mimic the sequestration time(Campbell & Kleckner, 1990).

To explore the dynamics of the oriC number following the cessation of dnaA expression, we can set the synthesis rate to be zero αA = 0. To simulate the dependence of initiation mass on dnaA expression, αA varies from 1hr−1ml−1 to 64hr−1ml−1.

Extrusion model

The extrusion model was formulated following the titration model, with an additional extruder introduced as a protein with constant concentration H = αHV with αH = 100 ml−1. The additional extruder competes the DnaA-boxes with DnaA proteins, with higher affinity to DnaA-boxes. So that it can displace DnaA from the DnaA-boxes, leading to a revised calculation for free DnaA concentration as [Af] = (A − (Ab − H))/V, under condition of (Ab − H) ≥ 0 and (A − (Ab − H)) ≥ 0.

To get the DnaA repression simulation at the population level, we introduced cell division by the assumption of Cooper and Helmstetter that a cell divides at a constant time (Tcyc = 60 mins) after DNA replication initiation. Meanwhile, we introduced noises on cell mass growth rate and the DnaA synthesis rate at each cell division: λ = λ0 + ηλ0, αA = α0 + ηα0 where λ0 = 0.66 hr−1, α0 = 5 hr−1ml−1 and η = 0.1 representing white noise. Symmetric cell division is used in the simulation for simplicity. This titration extrusion model is robust under various range of parameters.

DnaA-ATP/DnaA-ADP switch model

The switch model was simulated following the LDDR model in Berger et al. (Berger & Wolde, 2022). The same formulations of the DnaA-ATP fraction dynamics df/dt (equation 5) and the same parameters were used (Berger & Wolde, 2022). The cell division was determined when f reaches a threshold as defined previously (Berger & Wolde, 2022). To simulate the dnaA shut-down, we simply deleted the term of dnaA synthesis (λ(1 − f)).

Titration-switch model and titration-switch-extrusion model

In the integrated model combining titration and switch mechanisms, replication initiation is triggered when the concentration of free DnaA-ATP exceeds a threshold . The concentration of free DnaA-ATP is titrated by the DnaA-boxes on the chromosome as in the titration model, with , under condition of (AATP − Ab) ≥ 0. The parameters related to the titration dynamics remain consistent with those described in earlier sections of titration model.

The dynamics of total DnaA-ATP (AATP) is governed by both expression and conversion processes as described in Berger et al. (Berger & Wolde, 2022). According to this framework, the lipid and DARS systems facilitate the conversion of DnaA-ADP to DnaA-ATP at constant rates, while the datA site and RIDA systems also play a role in converting DnaA-ATP to DnaA-ADP, again at fixed rates. Additionally, the synthesis of new DnaA proteins provides another source of DnaA-ATP. For the parameters governing DnaA-ATP dynamics, we adhere to those specified by Berger et al., ensuring consistency with established research. To get a best fit of the initiation mass-dnaA expression relation, high threshold of free DnaA-ATP concentration was applied, .

To simulate the titration-switch-extrusion model, we incorporated the extruder as described in the extrusion model, where its amount was given by H = αHV. The concentration of free DnaA-ATP was then calculated as , under condition of (Abox − H) ≥ 0 and (AATP − (Ab − H)) ≥ 0. To explore the best fit of continuous DNA replication rounds following dnaA shut-down, αH = 400 ml−1 is used.

eLife Assessment

This work provides valuable insights by introducing a post-translational extrusion mechanism that could reshape how we understand the coupling between DnaA activity and DNA-replication initiation. While solid evidence is presented for some of the key results, other claims rest on indirect proxies and could be improved.

Reviewer #1 (Public review):

Summary:

The study by Li and coworkers addresses the important and fundamental question of replication initiation in Escherichia coli, which remains open, despite many classic and recent works. It leverages single-cell mRNA-FISH experiments in strains with titratable DnaA and novel DnaA activity reporters to monitor DNA activity peaks versus size. The authors find oscillations in DnaA activity and show that their peaks correlate well with the estimated population-average replication initiation volume across conditions and imposed dnaA transcription levels. The study also proposes a novel extrusion model where DNA-binding proteins regulate free DnaA availability in response to biomass-DNA imbalance. Experimental perturbations of H-NS support the model validity, addressing key gaps in current replication control frameworks.

Strengths:

I find the study interesting and well conducted, and I think its main strong points are:

(1) the novel reporters obtained with systematic synthetic biology methods, and combined with a titratable dnaA strain.

(2) the interesting perturbations (titration, production arrest, and H-NS).

(3) the use of single-cell mRNA FISH to monitor transcripts directly.

The proposed extrusion model is also interesting, though not fully validated, and I think it will contribute positively to the future debate.

Weaknesses and Limitations:

(1) A relevant limitation in novelty is that DnaA activity and concentration oscillations have been reported by the cited Iuliani and coworkers previously by dynamic microscopy, and to a smaller extent by the other cited study by Pountain and coworkers using mRNA FISH.

(2) An important limitation is that the study is not dynamic. While monitoring mRNA is interesting and relevant, the current study is based on concentrations and not time variations (or nascent mRNA). Conversely, the study by Iuliani and coworkers, while having the drawback of monitoring proteins, can directly assess production rates. It would be interesting for future studies or revisions to monitor the strains and reporters dynamically, as well as using (as a control) the technique of this study on the chromosomal reporters used by Iuliani et al.

(3) Regarding the mathematical models, a lot of details are missing regarding the definitions and the use of such models, which are only presented briefly in the Methods section. The reader is not given any tools to understand the predictions of different models, and no analytical estimates are used. The falsification procedures are not clear. More transparency and depth in the analysis are needed, unless the models are just used as a heuristic tool for qualitative arguments (but this would weaken the claims). The Berger model, for example, has many parameters and many regimes and behaviors. When models are compared to data (e.g., in Figure 2G), it is not clear which parameters were used, how they were fixed, and whether and how the model prediction depends on parameters.

(4) Importantly, the main statement about tight correlations of peak volumes and average estimated initiation volume does not establish coincidence, and some of the claims by the authors are unclear in these respects (e.g., when they say "we resolve a 1:1 coupling between DnaA activity thresholds and replication initiation", the statement could be correct but is ambiguous). Crucially, the data rely on average initiation volumes (on which there seems to be an eternally open debate, also involving the authors), and the estimate procedure relies on assumptions that could lead to biases and uncertainties added to the population variability (in any case, error bars are not provided).

(5) The delays observed by the authors (in both directions) between the peaks of DnaA-activity conditional averages with respect to volume and the average estimated initiation volumes are not incompatible with those observed dynamically by Iuliani and coworkers. The direct experiment to prove the authors' point would be to use a direct proxy of replication initiation, such as SeqA or DnaN, and monitor initiations and quantify DnaA activity peaks jointly, with dynamic measurements.

(6) While not being an expert, I had some doubt that the fact that the reporters are on plasmid (despite a normalization control that seems very sensible) might affect the measurements. Also, I did not understand how the authors validated the assumptions that the reporters are sensitive to DnaA-ATP specifically. It seems this assumption is validated by previous studies only.

Overall Appraisal:

In summary, this appears as a very interesting study, providing valuable data and a novel hypothesis, the extrusion model, open to future explorations. However, given several limitations, some of the claims appear overstated. Finally, the text contains some self-evaluations, such as "our findings redefine the paradigm for replication control", etc., that appear exaggerated.

Reviewer #2 (Public review):

Summary:

The authors show that in E. coli, the initiator protein DnaA oscillates post-translationally: its activity rises and peaks exactly when DNA replication begins, even if dnaA transcription is held constant. To explain this, they propose an "extrusion" mechanism in which nucleoid-associated proteins such as H-NS, whose amount grows with cell volume, dislodge DnaA from chromosomal binding sites; modelling and H-NS perturbations reproduce the observed drop in initiation mass and extra initiations seen after dnaA shut-down. Together, the data and model link biomass growth to replication timing through chromosome-driven, post-translational control of DnaA, filling gaps left by classic titration and ATP/ADP-switch models.

Strengths:

(1) Introduces an "extrusion" model that adds a new post-translational layer to replication control and explains data unexplained by classic titration or ATP/ADP-switch frameworks.

(2) A major asset of the study is that it bridges the longstanding gap between DnaA oscillations and DNA-replication initiation, providing direct single-cell evidence that pulses of DnaA activity peak exactly at the moment of initiation across multiple growth conditions and genetic perturbations.

(3) A tunable dnaA strain and targeted H-NS manipulations shift initiation mass exactly as the model predicts, giving model-driven validation across growth conditions.

(4) A purpose-built Psyn66 reporter combined with mRNA-FISH captures DnaA-activity pulses with cell-cycle resolution, providing direct, compelling data.

Weaknesses:

(1) What happens to the (C+D) period and initiation time as the dnaA mRNA level changes? This is not discussed in the text or figure and should be addressed.

(2) It is unclear what is meant by "relative dnaA mRNA level." Relative to what? Wild-type expression? Maximum expression? This should be explicitly defined.

(3) It would be helpful to provide some intuition for why an increase in dnaA mRNA level leads to a decrease in initiation mass per ori and an increase in oriC copy number.

(4) The titration and switch models do not explicitly include dnaA mRNA in the dynamics of DnaA protein. Yet, in Figure 2G, initiation mass is shown to decrease linearly with dnaA mRNA level in these models. How was dnaA mRNA level represented or approximated in these simulations?

(5) Is Schaechter's law (i.e., exponential scaling of average cell size with growth rate) still valid under the different dnaA mRNA expression conditions tested?

(6) The manuscript should explain more explicitly how the extrusion model implements post-translational control of DnaA and, in particular, how this yields the nonlinear drop in relative initiation mass versus dnaA mRNA seen in Figure 6E. Please provide the governing equation that links total DnaA, the volume-dependent "extruder" pool, and the threshold of free DnaA at initiation, and show - briefly but quantitatively - how this equation produces the observed concave curve.

(7) Does this Extrusion model give well well-known adder per origin, i.e., initiation to initiation is an adder.

(8) DnaA protein or activity is never measured; mRNA is treated as a linear proxy. Yet the authors' own narrative stresses post-translational (not transcriptional) control of DnaA. Without parallel immunoblots or activity readouts, it is impossible to know whether a six-fold mRNA increase truly yields a proportional rise in active DnaA.

(9) Figure 2 infers both initiation mass and oriC copy number from bulk measurements (OD₆₀₀ per cell and rifampicin-cephalexin run-out) instead of measuring them directly in single cells. Any DnaA-dependent changes in cell size, shape, or antibiotic permeability could skew these bulk proxies, so the plotted relationships may not accurately reflect true initiation events.

Author response:

The following is the authors’ response to the original reviews

Public Reviews:

Reviewer #1 (Public review):

Summary:

The study by Li and coworkers addresses the important and fundamental question of replication initiation in Escherichia coli, which remains open, despite many classic and recent works. It leverages single-cell mRNA-FISH experiments in strains with titratable DnaA and novel DnaA activity reporters to monitor DNA activity peaks versus size. The authors find oscillations in DnaA activity and show that their peaks correlate well with the estimated population-average replication initiation volume across conditions and imposed dnaA transcription levels. The study also proposes a novel extrusion model where DNA-binding proteins regulate free DnaA availability in response to biomass-DNA imbalance. Experimental perturbations of H-NS support the model validity, addressing key gaps in current replication control frameworks.

Strengths:

I find the study interesting and well conducted, and I think its main strong points are:

(1) the novel reporters obtained with systematic synthetic biology methods, and combined with a titratable dnaA strain.

(2) the interesting perturbations (titration, production arrest, and H-NS).

(3) the use of single-cell mRNA FISH to monitor transcripts directly.

The proposed extrusion model is also interesting, though not fully validated, and I think it will contribute positively to the future debate.

We thank the reviewer for acknowledging the strengths of our study.

Weaknesses and Limitations:

(1) A relevant limitation in novelty is that DnaA activity and concentration oscillations have been reported by the cited Iuliani and coworkers previously by dynamic microscopy, and to a smaller extent by the other cited study by Pountain and coworkers using mRNA FISH.

(2) An important limitation is that the study is not dynamic. While monitoring mRNA is interesting and relevant, the current study is based on concentrations and not time variations (or nascent mRNA). Conversely, the study by Iuliani and coworkers, while having the drawback of monitoring proteins, can directly assess production rates. It would be interesting for future studies or revisions to monitor the strains and reporters dynamically, as well as using (as a control) the technique of this study on the chromosomal reporters used by Iuliani et al.

We acknowledge the value of dynamic measurements and clarify our methodological rationale.

While luliani et al. provided valuable temporal resolution through protein dynamics, our mRNA FISH approach achieves direct decoupling of transcriptional vs. post-translational regulation (Fig 4F-H), and condition flexibility across 7 growth rates (30-66 min doubling times). This trade-off sacrifices temporal resolution for enhanced population-scale resolution and perturbation flexibility. To directly address temporal coupling, future work will implement dual-color live imaging of DnaA activity concurrent with replication initiation events.

(3) Regarding the mathematical models, a lot of details are missing regarding the definitions and the use of such models, which are only presented briefly in the Methods section. The reader is not given any tools to understand the predictions of different models, and no analytical estimates are used. The falsification procedures are not clear. More transparency and depth in the analysis are needed, unless the models are just used as a heuristic tool for qualitative arguments (but this would weaken the claims). The Berger model, for example, has many parameters and many regimes and behaviors. When models are compared to data (e.g., in Figure 2G), it is not clear which parameters were used, how they were fixed, and whether and how the model prediction depends on parameters.

We agree that model transparency is essential for quantitative validation. To address this, all model parameters (DnaA synthesis rate, activation/deactivation rates etc.) are explicitly tabulated in Supplementary Information Table S6. For the titration (Hansen et al. 1991) and extrusion models, we derive analytical expressions for initiation mass (IM) sensitivity to DnaA expression in Supplementary Note 1. For Figure 2G/S6, we used published parameters (Berger & Wolde 2022 SI Table 2) with experiment growth conditions (μ = 1.54 h-1).

The extrusion model's validation relies primarily on its ability to resolve paradoxical initiation events under dnaA shutdown (Fig 6C), a test where other models fail categorically. While the Berger titration-switch hybrid can fit steady-state IM trends (Fig S6A), it cannot reproduce post-shutdown dynamics without ad hoc modifications (Fig S6B). We acknowledge that comprehensive analysis of all model regimes exceeds this study's scope but provide full simulation code for independent verification: https://github.com/BaiYangBqdq/dynamics_of_biomass_DNA_coordination

(4) Importantly, the main statement about tight correlations of peak volumes and average estimated initiation volume does not establish coincidence, and some of the claims by the authors are unclear in these respects (e.g., when they say "we resolve a 1:1 coupling between DnaA activity thresholds and replication initiation", the statement could be correct but is ambiguous). Crucially, the data rely on average initiation volumes (on which there seems to be an eternally open debate, also involving the authors), and the estimate procedure relies on assumptions that could lead to biases and uncertainties added to the population variability (in any case, error bars are not provided).

We acknowledge the limitations of population-level inference and have refined our claims: "Replication initiation volume scales proportionally with peak DnaA activity volume with a slope of 1.0 (R2=0.98, Fig 7G), indicating predictive correspondence rather than absolute coincidence. While population-level 𝑉𝑖 estimation cannot resolve single-cell stochasticity, the consistent 𝑉*: 𝑉𝑖 relationship across 20 conditions suggest DnaA activity thresholds predict initiation timing within physiological error margins”. Future work will implement simultaneously DnaA activity and replication forks by using microfluidic single-cell tracking.

(5) The delays observed by the authors (in both directions) between the peaks of DnaAactivity conditional averages with respect to volume and the average estimated initiation volumes are not incompatible with those observed dynamically by Iuliani and coworkers. The direct experiment to prove the authors' point would be to use a direct proxy of replication initiation, such as SeqA or DnaN, and monitor initiations and quantify DnaA activity peaks jointly, with dynamic measurements.

We acknowledge the observed temporal deviations between DnaA activity peaks (𝑉*) and population-derived volumes at initiation ( 𝑉𝑖) in certain conditions, in line with the findings of Iuliani et al. This might be mechanistically consistent with the time required for orisome assembly or oriC sequestration. They do not contradict our core finding that initiation occurs at a defined DnaA activity threshold (slope=1.0, R2=0.98 in 𝑉*: 𝑉𝑖 correlation).

(6) While not being an expert, I had some doubt that the fact that the reporters are on plasmid (despite a normalization control that seems very sensible) might affect the measurements. Also, I did not understand how the authors validated the assumptions that the reporters are sensitive to DnaA-ATP specifically. It seems this assumption is validated by previous studies only.

We employed a plasmid-based reporter system to circumvent the significant confounding effects of chromosomal position on promoter activity, as extensively documented by Pountain et al., where local genomic context (e.g., nucleoid occlusion, supercoiling gradients, and neighboring operons) introduces uncontrolled variability. By housing the Psyn66 test promoter and Pcon normalization control in identical low-copy pSC101 vectors (<8 copies/ cell, Peterson & Phillips, Plasmid 2008), we ensured they experience equivalent physical and biochemical environments. This ratiometric design, where DnaA activity is calculated, actively corrects for global fluctuations in RNA polymerase availability, nucleotide pools, and plasmid copy number. Critically, Psyn66’s architecture emulates natural DnaA-responsive elements: its strong DnaAboxes report free DnaA concentration, while its weak box is preferentially bound by DnaA-ATP (Speck et al., EMBO journal 1999), mirroring the nucleotide-state sensitivity of oriC and the native dnaA promoter. This system was indispensable for our central finding, as it uniquely enabled the decoupling of DnaA activity oscillations from transcriptional feedback (Fig. 4F-H), an experiment fundamentally impossible with chromosomally integrated reporters due to autoregulatory interference.

Overall Appraisal:

In summary, this appears as a very interesting study, providing valuable data and a novel hypothesis, the extrusion model, open to future explorations. However, given several limitations, some of the claims appear overstated. Finally, the text contains some selfevaluations, such as "our findings redefine the paradigm for replication control", etc., that appear exaggerated.

We thank the reviewer for highlighting the need for precise language in framing our conclusions. We have implemented the following substantive revisions throughout the manuscript to ensure claims align strictly with empirical evidence:

(1) Changed "redefine the paradigm for replication control" into "advance the paradigm for replication control" (Introduction)

(2) Changed "redefine bacterial cell cycle control" into "refine bacterial cell cycle control as a dynamic interplay..." (Discussion)

(3) Removed the term "spatial" from the Discussion's description of DnaA-chromosome interactions (Discussion, first paragraph).

(4) Changed "provides a blueprint" into "provides a valuable tool for dissecting spatial regulation..." (Discussion, final paragraph)

(5) Scrutinized all superlatives (e.g., "critical feat" into "important capability"; "fundamental principle of cellular organization" into "potential organizational strategy")

(6) Replaced the instances of "robust" with evidence-backed descriptors (e.g., "sensitive," "consistent")

(7) We agree that the extrusion model requires further validation and have emphasized this in Discussion: "While H-NS perturbation supports extrusion mechanism, future work should identify the full extruder interactome and elucidate how metabolic signals modulate their activity" (final paragraph)

This calibrated language more accurately represents our study as a conceptual advance with testable mechanisms, not a complete paradigm shift.

Reviewer #2 (Public review):

Summary:

The authors show that in E. coli, the initiator protein DnaA oscillates post-translationally: its activity rises and peaks exactly when DNA replication begins, even if dnaA transcription is held constant. To explain this, they propose an "extrusion" mechanism in which nucleoidassociated proteins such as H-NS, whose amount grows with cell volume, dislodge DnaA from chromosomal binding sites; modelling and H-NS perturbations reproduce the observed drop in initiation mass and extra initiations seen after dnaA shut-down. Together, the data and model link biomass growth to replication timing through chromosome-driven, posttranslational control of DnaA, filling gaps left by classic titration and ATP/ADP-switch models.

Strengths:

(1) Introduces an "extrusion" model that adds a new post-translational layer to replication control and explains data unexplained by classic titration or ATP/ADP-switch frameworks.

(2) A major asset of the study is that it bridges the longstanding gap between DnaA oscillations and DNA-replication initiation, providing direct single-cell evidence that pulses of DnaA activity peak exactly at the moment of initiation across multiple growth conditions and genetic perturbations.

(3) A tunable dnaA strain and targeted H-NS manipulations shift initiation mass exactly as the model predicts, giving model-driven validation across growth conditions.

(4) A purpose-built Psyn66 reporter combined with mRNA-FISH captures DnaA-activity pulses with cell-cycle resolution, providing direct, compelling data.

We thank the reviewer for acknowledging the strengths of our study.

Weaknesses:

(1) What happens to the (C+D) period and initiation time as the dnaA mRNA level changes? This is not discussed in the text or figure and should be addressed.

We thank the reviewer for this important observation. Our data demonstrate that increased dnaA mRNA levels induce two compensatory changes in cell cycle progression:

(1) Earlier replication initiation, manifested as a reduced initiation mass: the initiation mass decreased from 5.6 to 2.6 (OD600·ml per 1010 cells) as the relative dnaA mRNA level increased from 0.2 to 7.2 (normalized to the wild-type level) (Fig. 2F, red).

(2) Prolonged C+D period: Increased by approximately 60% (from 1.05 to 1.66 hours, Fig. 2F blue).

The complete quantitative relationship is now explicitly described in the Results section: “Concurrently, the initiation mass was reduced by 50%, and the period from initiation to division (C+D) was increased by ~60% (Fig. 2F)”

(2) It is unclear what is meant by "relative dnaA mRNA level." Relative to what? Wild-type expression? Maximum expression? This should be explicitly defined.

The relative dnaA mRNA level was obtained by normalizing to that in wild-type MG1655 cells grown in the same medium. To clarify this point, we have now marked the wild-type level in Fig. 1B, and a clear description of this has also been included in the figure caption.

(3) It would be helpful to provide some intuition for why an increase in dnaA mRNA level leads to a decrease in initiation mass per ori and an increase in oriC copy number.

Thank you for your valuable suggestion. Increased dnaA mRNA accelerates DnaA accumulation, causing cells to reach the initiation threshold at a smaller cell size (reducing initiation mass, Fig. 2F red). This earlier initiation increases oriC copies per cell at populational level (Fig. 2E). This mechanistic interpretation now appears in the Results: “As the DnaA expression level increases, DnaA activity reaches the initiation threshold earlier. Given that cell mass remained nearly unchanged, this earlier initiation led to an increase in population-averaged cellular oriC numbers (Fig. 2E).”

(4) The titration and switch models do not explicitly include dnaA mRNA in the dynamics of DnaA protein. Yet, in Figure 2G, initiation mass is shown to decrease linearly with dnaA mRNA level in these models. How was dnaA mRNA level represented or approximated in these simulations?

All models presented in this article omit explicit modeling of dnaA mRNA dynamics for simplicity. However, at steady state, the relative level of dnaA mRNA can be approximated by the relative expression rate of DnaA protein, as both reflect the expression level of DnaA. This detail is now clarified in the caption of Figure 2G.

(5) Is Schaechter's law (i.e., exponential scaling of average cell size with growth rate) still valid under the different dnaA mRNA expression conditions tested?

Schaechter's law describes the exponential scaling of average cell size with growth rate in bacteria. In our prior work (Zheng et al., Nature Microbiology 2020), where we demonstrated that Schaechter's law fails in slow-growth regimes. However, in current study, growth rate remained constant across different dnaA expression levels (Fig. 2C), and cell mass showed no significant change (Fig. 2D). Since Schaechter's law specifically addresses how cell size scales with growth rate, it does not apply here, as growth rate was invariant in our perturbations, which selectively alter replication initiation dynamics, not growth rate or size scaling.

(6) The manuscript should explain more explicitly how the extrusion model implements posttranslational control of DnaA and, in particular, how this yields the nonlinear drop in relative initiation mass versus dnaA mRNA seen in Figure 6E. Please provide the governing equation that links total DnaA, the volume-dependent "extruder" pool, and the threshold of free DnaA at initiation, and show - briefly but quantitatively - how this equation produces the observed concave curve.

The governing equations linking initiation mass and DnaA expression level is now provided in Supplementary Note S1 for both the titration and the extrusion model. In general, the dependence of initiation mass (𝑉𝐼) on dnaA expression level (𝛼𝐴) dependency takes an inverse 1 proportionality form: . In the extrusion model, the incorporated extruder protein is assumed to have similar synthesis dynamics as DnaA and can release DnaA from DnaA-box. After denoting the synthesis rate of the extruder as 𝛼𝐻, the combined effect of DnaA and the extruder on replication initiation can be briefly described as: . Then the additive contribution of 𝛼𝐻 dampens the sensitivity of initiation mass to changes in 𝛼𝐴, resulting in a significantly flattened curve. As a result, the predicted 𝑉𝐼 − 𝛼𝐴 relationship has a concave shape in the semi-log plots.

(7) Does this Extrusion model give well well-known adder per origin, i.e., initiation to initiation is an adder.

Yes, the extrusion model can provide the initiation-to-initiation adder phenomenon, this information was provided in fig. S3C.

(8) DnaA protein or activity is never measured; mRNA is treated as a linear proxy. Yet the authors' own narrative stresses post-translational (not transcriptional) control of DnaA. Without parallel immunoblots or activity readouts, it is impossible to know whether a sixfold mRNA increase truly yields a proportional rise in active DnaA.

We acknowledge the reviewer's valid concern regarding the indirect nature of our DnaA activity measurements. While mRNA levels alone cannot resolve active DnaA dynamics, our approach integrates functional replication outcomes with a validated synthetic reporter to infer activity. Crucially, elevated dnaA mRNA causes demonstrable biological effects: earlier replication initiation (Fig. 2F) and increased oriC copies (Fig. 2E), directly confirming enhanced functional DnaA activity at the oriC locus. The Psyn66 reporter, engineered with DnaA-boxes mirroring oriC's architecture, provides orthogonal validation, showing progressive repression to dnaA induction (Fig. 3C). Our operational metric , bases on Psyn66 responds sensitively to DnaA-chromosome interactions within its characterized 8-fold dynamic range (Fig. 3C). Immunoblots would be inadequate here, as they cannot distinguish functionally critical pools: free versus chromosome-bound DnaA, or DnaA-ATP versus DnaAADP, precisely the post-translational states our study implicates in regulation. We therefore prioritize functional readouts (initiation timing) and the Psyn66 reporter, which probes the biologically active fraction relevant to replication control.

(9) Figure 2 infers both initiation mass and oriC copy number from bulk measurements (OD600 per cell and rifampicin-cephalexin run-out) instead of measuring them directly in single cells. Any DnaA-dependent changes in cell size, shape, or antibiotic permeability could skew these bulk proxies, so the plotted relationships may not accurately reflect true initiation events.

We acknowledge the reviewer's valid methodological concern and clarify that while bulk measurements carry inherent limitations, our approach is grounded in established techniques with demonstrated reliability. Cell mass was inferred from OD600/cell, which correlates strongly with direct dry weight measurements and microscopic cell volumes across diverse growth conditions, as validated in our prior work (Zheng et al., Nature Microbiology 2020). Crucially, cell mass remained invariant across dnaA expression levels (Fig. 2D).

Regarding oriC quantification, the rifampicin-cephalexin run-out assay is a wildly applied for replication initiation studies. Our data shows expected 2n oriC distributions without abnormal ploidy (as shown below). While single-cell methods offer superior resolution, our bulk approach provides accurate population-level trends.

Recommendations for the authors:

Reviewing Editor Comments:

The reviewers felt that the mathematical modeling was not adequately explained in the paper, and that this affected the readability of the manuscript. The authors are encouraged to elaborate on this aspect of the paper (in addition to strengthening other claims, if possible, per the reviewers' comments).

We thank the editor and reviewers for their constructive feedback. We have comprehensively strengthened the mathematical modeling framework to enhance clarity and rigor.

Reviewer #1 (Recommendations for the authors):

The only revision I would do is a recalibration of the claims and a major effort to clarify the modeling part (including a detailed SI appendix), without necessarily performing additional work.

To enhance mathematical modeling transparency, we have completed model description in the method section and a parameter table with literature-sourced values in Supplementary Information Table S6. Moreover, analytical derivations of initiation mass dependencies are performed and presented in the Supplementary Information Note S1.

Of course, there are extra experiments (mentioned in the public review) that would help support some of the big claims, but that can be considered a different project.

Thank you for your suggestion. This will be addressed in our future work.

Minor suggestion: please put signposts or plot jointly to compare the maxima/minima in Figures 4D, E, G, and H.

We added dashed lines in Figures 4D, and E, to synchronize visualization of DnaA activity peaks and transcriptional minima across panels, facilitating direct biological comparisons.

Reviewer #2 (Recommendations for the authors):

(1) Should define what DNA activity is.

We have explicitly defined DnaA activity in the Introduction as “the capacity to initiate replication…” and noted that it is “governed by free DnaA concentration, DnaA-ATP/-ADP ratio, and orisome assembly competence”.

(2) Word repetition - “...grown in in Luria-Bertani (LB) medium...”.

Corrected.

(3) Typographical error - “FISH ... was preformed" should be "performed”.

Corrected.

(4) The manuscript alternates between “ng ml-1” and “ng·ml-1”; choose one style and apply it uniformly.

Standardized the units to ng·ml-1 throughout.

(5) Reference duplicates - Some citations appear twice in the bibliography (e.g., "Bintu et al., 2005a/b" and "Bintu et al., 2005b" listed again later).

The studies by Bintu et al. (2005a, 2005b) represent separate works: 2005a details applications, and 2005b develops models.

Version 2

Abstract

Abstract

Robust control of DNA replication is fundamental to bacterial proliferation. In Escherichia coli, replication initiation is thought to be regulated by oscillations in DnaA activity, driven by DnaA-chromosome interactions that differ among leading models. However, direct evidence linking these oscillations to replication initiation has been lacking, and existing models fail to explain the observed decoupling of replication initiation from dnaA expression. Here, we establish a direct link between DnaA activity and replication initiation by demonstrating robust oscillations in DnaA activity, which peak precisely at replication initiation across diverse growth conditions and genetic perturbations. Notably, these oscillations persist even when dnaA transcription remains constant, suggesting a regulatory mechanism that modulates DnaA activity independently of its expression. Additionally, we propose an extrusion model in which DNA-binding proteins sense biomass-DNA imbalance and extrude DnaA from the chromosome to trigger replication, overcoming limitations of existing models. Consistent with this model, perturbation of the nucleoid-associated protein H-NS modulates DnaA activity and replication timing, supporting its mechanistic validity.

Introduction

Bacterial proliferation requires precise coordination between DNA replication and biomass growth to maintain cellular homeostasis. While biomass accumulates exponentially, DNA synthesis proceeds in a stepwise manner, creating temporal imbalances that demand continuous correction throughout the cell cycle (Fig. 1A-B). Failure to synchronize these processes risks catastrophic outcomes, such as growth defects, incomplete replication, unequal genetic segregation, and cell death (Boye et al., 1996; Khodursky et al., 2015; Lu et al., 1994; Mäkelä et al., 2024; Zaritsky & Pritchard, 1971). Central to this synchronization in Escherichia coli is the initiator protein DnaA, a DNA-binding protein that binds specifically to an asymmetric 9-bp DnaA-box, whose consensus sequence is TTWTNCACA (Hansen & Atlung, 2018; Speck et al., 1999). Although DnaA binding to the replication origin (oriC) triggers replication initiation (Dong et al., 2023; Kaguni, 2011; Ohbayashi et al., 2020; Ozaki & Katayama, 2012; Sekimizu et al., 1987), how DnaA activity—defined as the capacity to initiate replication—is dynamically regulated to coordinate DNA synthesis with cellular growth, particularly when DNA content and biomass diverge, remains a fundamental unresolved question in cell cycle biology.

Coordination of biomass growth and DNA replication through replication initiation and DnaA activity oscillations.

(A) Schematic representation of the discrepancy between biomass accumulation and DNA replication. While biomass grows exponentially, DNA synthesis progresses linearly, necessitating replication initiation events to maintain coordination. (B) Mechanistic model for biomass-DNA coordination in bacteria. A molecular sensor detects deviations between cell mass and DNA content, transmitting this information to regulatory controllers that compensate by either increasing DNA replication or restricting biomass accumulation. (C) Illustration of cyclic DnaA activity oscillations aligning with replication initiation to ensure precise cell cycle control.

Two competing models have dominated the field. The titration model posits that DnaA activity is governed by the availability of free DnaA, which accumulates until it surpasses the buffering capacity of chromosomal DnaA-boxes (Christensen et al., 1999; Hansen et al., 1991). Here, replication initiation resets the system by increasing DNA content (and thus DnaA-boxes), titrating excess DnaA. In contrast, the switch model emphasizes the ATP/ADP-bound states of DnaA: DnaA-ATP, synthesized during biomass growth, triggers initiation, while hydrolysis to DnaA-ADP following replication resets the cycle (Berger & Wolde, 2022; Donachie & Blakely, 2003; Fu et al., 2023; Hansen & Atlung, 2018; Katayama et al., 2017). Crucially, the titration model depends on ongoing DnaA synthesis to reflect biomass accumulation, whereas the switch model relies on nucleotide-state transitions, independent of total DnaA levels. Despite these mechanistic differences, both models consider that replication initiation is regulated by oscillations in DnaA activity, driven by DnaA-chromosome interactions (Fig. 1C). However, direct evidence linking these oscillations to replication initiation remains lacking. Moreover, neither model fully explains recent experimental observations: replication initiation persists for multiple generations even when dnaA transcription is completely inhibited (Knöppel et al., 2023). This discrepancy suggests an uncharacterized mechanism that senses biomass-DNA imbalance and directly modulates DnaA activity.

Here, using a synthetic reporter coupled with single-cell mRNA fluorescence in situ hybridization (FISH), we demonstrate that DnaA activity—governed by free DnaA concentration, DnaA-ATP/-ADP ratio, and orisome assembly competence—peaks precisely at initiation across diverse growth conditions and genetic backgrounds, even when dnaA transcription is held constant. DnaA activity oscillations persist independently of transcriptional feedback, pointing to a post-translational regulatory mechanism. We propose an extrusion model in which nucleoid-associated proteins (NAPs) sense biomass-DNA imbalance and dynamically extrude DnaA from the chromosome, liberating it to activate oriC. This model not only explains the decoupling of replication timing from dnaA expression but also integrates disparate experimental findings. Perturbations of the NAP, H-NS for instance, modulate DnaA activity and replication timing, directly supporting the model’s predictions. By unifying regulatory principles across transcription, protein activity, and chromosome dynamics, our findings advance the paradigm for bacterial replication control.

Results

Development of a DnaA activity reporter system

It has been known that changing DnaA expression levels affects DnaA activity, and further changes initiation mass (Berger & Wolde, 2022; Løbner-Olesen et al., 1989; Zheng et al., 2020). To investigate this further, we engineered a dnaA-titratable strain, where dnaA transcription is controlled by an inducible Ptet promoter (Lutz & Bujard, 1997) (Fig. 2A). This system enabled precise modulation of dnaA mRNA levels (0.25 to 6 times the wild-type level) via anhydrotetracycline (aTc) induction (0.75– 20 ng·ml−1) without altering growth rate or cell size (Fig. 2B–D). As the DnaA expression level increases, DnaA activity reaches the initiation threshold earlier. Given that cell mass remained nearly unchanged, this earlier initiation led to an increase in population-averaged cellular oriC numbers (Fig. 2E). Concurrently, the initiation mass was reduced by 50%, and the period from initiation to division (C + D) was increased by ∼60% (Fig. 2F). Notably, the observed relationship between DnaA expression and initiation mass deviated significantly from both the titration and switch models (Fig. 2G; Supplementary Note S1), suggesting unaccounted regulatory mechanisms.

Construction and characterization of the dnaA-titratable strain.

(A) Schematic of the dnaA-titratable strain. A dnaA gene under the control of the Ptet promoter was inserted near oriC and regulated by a Ptet-tetR feedback loop integrated at the intS locus, enabling fine-tuned expression control. The native dnaA gene was replaced with a kanamycin resistance cassette (kanr). (B–F) Characterization of dnaA-titratable cells (circle) and wild-type MG1655 cells (triangle) grown in rich defined medium with glycerol (M6) under varying aTc concentrations. Measured parameters include: (B) dnaA mRNA levels; (C) growth rate; (D) population-averaged cellular mass; (E) population-averaged oriC numbers; (F) initiation mass (red, left axis) and the initiation to division period (C+D) (blue, right axis). The dnaA mRNA levels were normalized to that in wild-type cells. Cellular mass was determined by OD6000 divided by cell number concentration. oriC copy numbers were measured using a run-out assay, and initiation mass was calculated as the ratio of cellular mass to oriC numbers. Data represent means ± SD (n = 5 biological replicates). (G) Relationship between relative initiation mass and relative dnaA mRNA levels, compared with predictions from the initiation titration model (blue line) and the switch model (purple line). The relative dnaA mRNA levels in experiments are compared to relative DnaA expression rate αA in models. Experimental data are overlaid for validation.

To dissect DnaA activity dynamics, we constructed a library of synthetic promoters (n = 67) by replacing tetO operators in Ptet with DnaA-boxes (Fig. S1). To assess the degree of DnaA activity-mediated repression, we measured the fold change in GFP expression driven by each synthetic promoter under low and high levels of dnaA induction (Fig. 3A). Six synthetic promoters exhibited over 8-fold DnaA-mediated repression, whereas the promoter lacking DnaA-boxes (Pcon) showed none, and the well-characterized native dnaA promoter (Pnative)(Braun et al., 1985; Saggioro et al., 2013; Speck et al., 1999) displayed approximately two-fold repression (Fig. 3B; Fig. S1). Psyn66, which contains three strong and one weak DnaA-box, exhibited >8-fold repression under high DnaA induction (Fig. 3B). Psyn66’s design enables responsiveness to both free DnaA levels (via strong boxes) and the DnaA-ATP/DnaA-ADP ratio (via differential binding to the weak box)(Grimwade & Leonard, 2021; Katayama et al., 2017; Speck et al., 1999). Dose-response assays confirmed that Psyn66 was progressively repressed across a 40-fold range of dnaA expression (Fig. 3C), with no interference from SeqA-mediated sequestration (Fig. 3D-E). These results validated Psyn66 as a sensitive and specific reporter for DnaA activity.

Development of a DnaA activity reporter system.

(A) Schematic of promoter design and screening. Sixty-seven synthetic promoters were constructed by inserting various DnaA-boxes around the promoter core to drive gfp expression in a dnaA-titratable strain, where dnaA expression was regulated by aTc concentration. After pre-cultivation, GFP fluorescence per OD600 was measured using a microplate reader in cells grown under low (0.5 ng·ml−1) or high (50 ng·ml−1) aTc concentrations to determine repression fold-change. (B) Repression fold-change of synthetic promoters. Pcon (a promoter lacking DnaA-boxes) and Pnative (the endogenous dnaA promoter) served as negative and positive controls, shown in gray and yellow, respectively. (C) Response curves of three promoters to varying dnaA expression levels, with their promoter architectures shown on the right. Promoter activity was assessed by relative gfp mRNA levels, normalized to the lowest dnaA expression condition. Schematic (D) and response curves (E) of Psyn66 and Pnative responses to SeqA in the seqA-titratable strain. Promoter activity was quantified from gfp transcript levels in seqA-titratable cells containing the Psyn66-GFP plasmid, normalized to the lowest seqA expression level. All cells were grown in rich defined medium supplemented with glycerol across different aTc concentrations. Data represent mean ± SD from 3 biological replicates (B,C,E).

Decoupling DnaA activity oscillations from dnaA transcription fluctuations

To investigate DnaA activity dynamics throughout the cell cycle, we linked Psyn66 activity to cell size, a surrogate for cell-cycle progression, using single-cell mRNA FISH(Skinner et al., 2013) to quantify lacZ mRNA levels in MG1655Δlac cells harboring the Psyn66-lacZ plasmid (Fig. 4A). A control strain harboring a promoter-less lacZ construct (Pneg) showed undetectable fluorescence signals (Fig. 4A), confirming the specificity of lacZ mRNA detection. Quantitative analysis revealed that Psyn66-driven mRNA levels fluctuated approximately 3-fold over the cell cycle, whereas the DnaA-unresponsive constitutive promoter (Pcon) exhibited stable expression (Fig. 4B). However, fluctuations in mRNA concentration may not directly correspond to changes in DnaA activity due to global factors such as plasmid copy number and RNA polymerase concentration (Balakrishnan et al., 2022; Bintu et al., 2005a; Bintu et al., 2005b). To account for these variables, we used a Pcon-lacZ plasmid to estimate global effects on mRNA concentration and calculated DnaA activity by assuming that it is inversely proportional to . In wild-type cells (MG1655Δlac) harboring the synthetic reporter system (Fig. 4C), fluctuations in DnaA activity, as denoted by , displayed approximately a 3-fold variation across the cell cycle (Fig. 4D), underscoring the substantial cell cycle-dependent oscillations in DnaA activity.

DnaA activity oscillations decoupled from dnaA transcription fluctuations.

(A) Representative lacZ mRNA FISH images of MG1655 Δlac cells transformed with lacZ expression plasmids driven by Psyn66 (DnaA-boxes around promoter core), Pcon (no DnaA-box around promoter core), or Pneg (mutated promoter core). Yellow outlines indicate cell boundaries identified from phase-contrast images. (B) Relative lacZ mRNA concentrations driven by Psyn66 (left) and Pcon (right) across different cell volumes. Relative concentrations were determined from volume-specific lacZ mRNA fluorescence intensities, normalized to the population average. Volume-binned data for Psyn66 ([mZ] (Psyn66)) and Pcon ([mZ] (Pcon)) are shown as open circles and were used to calculate ksyn66. (C) Schematic of a strain with autoregulated dnaA transcription carrying a DnaA activity reporter plasmid. Cell cycle-dependent fluctuations in relative DnaA activity (D) and relative dnaA mRNA concentrations (E) in cells from panel C, grown in rich defined medium supplemented with glucose. Relative DnaA activity , calculated from volume-binned data in panel B, was smoothed and plotted as a red curve (D). Relative dnaA mRNA concentrations were determined from volume-specific dnaA mRNA fluorescence intensities, normalized to the population average, with volume-binned data shown as open circles (E). Dashed lines indicate the cell volume at peak DnaA activity (D) and the minimum dnaA mRNA content (E). (F–H) Same as panels C–E, but for cells with aTc induced dnaA transcription, grown in rich defined medium supplemented with glycerol under 2 ng·ml−1 aTc induction. More than 8,000 cells were analyzed per growth condition, and all error bars correspond to standard error of the mean (SEM).

We then examined whether the observed fluctuations in DnaA activity were linked to oscillations in dnaA transcription. Comparing the dynamics of dnaA mRNA and DnaA activity, we found that the peak in DnaA activity coincided with the trough in dnaA mRNA levels (dashed line in Fig. 4D-E). To test whether dnaA transcription drives these oscillations, we eliminated transcriptional fluctuations using the dnaA-titratable strain. Despite constant dnaA mRNA levels (2 ng·ml−1 aTc induction), DnaA activity was retained consistent oscillations (Fig. 4F–H), demonstrating the decoupling of DnaA activity oscillations from transcriptional fluctuations. Our data suggest that post-translational regulation, not transcriptional feedback, governs the DnaA activity.

DnaA activity oscillations are tightly coupled to replication timing

To explore the relationship between DnaA activity oscillations and DNA replication initiation, we examined cells grown under various nutrient conditions supporting different doubling times (30–66 min). Significant fluctuations in DnaA activity were observed across all conditions (Fig. 5A). We also estimated the cell volume at replication initiation (Vi) based on population-averaged cell volume and oriC number (Bremer et al., 1979; Si et al., 2017; Zheng et al., 2016) (Fig. S2 A–C), marking this on the DnaA activity profiles. In a representative birth-to-division cell cycle (Pountain et al., 2024) (Fig. S2D), DnaA activity peaks consistently coincided with Vi, indicating a close correlation between DnaA activity and replication initiation (Fig. 5A).

Tight correlation between DnaA activity oscillations and DNA replication initiation.

(A) Cell cycle-dependent DnaA activity oscillations in wild-type cells across various growth conditions. DnaA activity is represented by . Volume-binned DnaA activity (red circles) was smoothed and plotted as a red curve. The representative birth-to-division cell cycle, defined as the cell volume doubling interval containing the majority of cells, is shaded in gray. The vertical line indicates the cell volume at replication initiation (Vi). (B) Cell cycle-dependent DnaA activity oscillations in dnaA-titratable cells grown in M6 medium with varying aTc concentrations. More than 8,000 cells were analyzed per growth condition, and error bars show the mean ± SEM. (C) Correlation between V∗ and Vi in wild-type (squares) and dnaA-titratable (circles) cells. V∗ represents the cell volume at the peak of DnaA activity within the representative birth-to-division cell cycle. The black line indicates equivalence between V∗ and Vi.

Further genetic perturbations, such as titrating DnaA expression, shifted DnaA activity fluctuations, yet the relationship between V∗ (the cell volume at maximal DnaA activity in a representative birth-to-division cell cycle) and Vi remained locked (Fig. 5B). To quantify this relationship, we compared Vi with V∗, revealing a strictly proportional relationship through the origin (slope= 1.0, R2 = 0.98), indicating equivalence between V∗ and Vi (Fig. 5C). These results confirm that DnaA activity oscillations are tightly correlated with replication initiation, independent of transcriptional or nutrient perturbations.

An extrusion model reconciles paradoxical initiation events

Very recently, Knöppel et al. reported that multiple rounds of initiation occurred even after DnaA synthesis is shut down (Knöppel et al., 2023). In their experiment, DnaA was supplied in excess before expression ceased (Knöppel et al., 2023), potentially allowing continued replication initiation due to residual DnaA. To address this issue, we shut down DnaA expression from its native level using the deactivated miniature Un1Cas12f1 (dUn1Cas12f1) system, guided by sgRNA to repress dnaA transcription (Fig. 6A). This system effectively diminished dnaA transcription to an extremely low level, yet total oriC numbers increased fourfold within 90 min, consistent with two rounds of replication initiation (Fig. 6B). These findings demonstrate that halting DnaA synthesis does not immediately abolish replication initiation.

An extrusion model explains DnaA shut-down dynamics.

(A) Genetic circuit of the deactivated CRISPR-Cas system for dnaA transcription shut-down. dnaA gene is targeted by a constitutively expressed sgRNA, while dUn1Cas12f1 expression is inhibited by TetR repressor. These transcription units are separated by terminators. The cassette was integrated into the chromosome near the oriC locus. DnaA shutdown is induced by the addition of aTc. (B) Time-course of relative dnaA mRNA levels (red line, left axis) and total oriC number (green line, right axis) following the addition of 50 ng·ml−1 aTc at time 0 (dashed line). dnaA mRNA levels were normalized to wild-type levels, and oriC numbers were normalized to their initial values. Error bars indicate mean ± SD (n = 3 biologically independent experiments). (C) Predicted increases in total oriC number during dnaA transcription shut-down based on three models: the titration model, switch model, and extrusion model. Shutdown was simulated by setting dnaA transcription to zero at time 0 (dashed line). (D) Schematic of the extrusion model. The model introduces extruder(s) as additional regulators of biomass-DNA coordination, complementing the role of DnaA (left). Increased binding of the extruder to DNA promotes the release of DnaA from DnaA-boxes (right). (E) Comparison of the relationship between relative initiation mass and relative dnaA mRNA levels from experimental data (Fig. 2F) and predictions of the extrusion model.

Both the titration and switch models predict a close relationship between DnaA activity oscillations and replication initiation(Berger & Wolde, 2022). However, they fail to account for the multiple rounds of initiation observed after DnaA synthesis was shut down. According to the titration model, initiation should cease immediately once DnaA expression is shut down, as the reduction in DnaA concentration would prevent it from reaching the threshold required to trigger new rounds of replication. In contrast, the switch model predicts infinite initiations following the cessation of DnaA expression (Fig. 6C). This is because the cessation of DnaA synthesis only reduce the production rate of DnaA-ATP and had no effect on the conversion rates between DnaA-ATP and DnaA-ADP. As a result, the ratio of DnaA-ATP to DnaA-ADP would temporarily decrease, causing a brief lag before initiation resumes indefinitely (Fig. 6C).

The failure of both models prompts us to revisit the regulatory mechanisms of replication initiation. In the titration model, a substantial fraction of DnaA proteins binds to DnaA-boxes along the chromosome, raising the possibility that these bound proteins serve as a reservoir of free DnaA, sustaining replication initiation even after DnaA synthesis ceases. We thus considered the possibility of some extruder(s) alongside DnaA to sense biomass-DNA deviation, and flush DnaA proteins away from the DnaA-boxes, thus releasing free DnaA to dynamically restore DnaA activity, even in the absence of ongoing DnaA synthesis (Fig. 6D).

In this proposed “extrusion model”, the extruder is constitutively expressed, synthesized at a constant rate throughout the cell cycle, accumulating in proportion to biomass growth, enabling replication initiation when free DnaA concentrations exceed the critical threshold, as described in the titration model(Hansen & Atlung, 2018; Hansen et al., 1991) (Fig. 6D). This model guarantees the long-term coordination of biomass growth and oriC abundance, as well as stable oscillations in free DnaA and oriC concentrations (Fig. S3 A). Notably, the extrusion model accounts for the persistence of replication initiation and free DnaA oscillations over multiple cell cycles after DnaA synthesis is halted (Fig. 6C; Fig. S3 B). Moreover, since the extruder mitigates the impact of perturbations on DnaA synthesis, the model predicts a quantitative relationship between initiation mass and DnaA expression levels that aligns with experimental observations (Fig. 6E; Supplementary Note S1). Furthermore, a stochastic implementation of the extrusion model shows the lack of correlation between successive replication initiation events (Fig. S3 C), as observed in recent studies (Si et al., 2019; Witz et al., 2019).

Perturbation of H-NS validates the extrusion mechanism

The extrusion model could be validated by changing the expression level of the extruder, which modulates DnaA activity by releasing DnaA from DnaA-boxes. Given their DNA binding properties, NAPs(Dillon & Dorman, 2010) are likely candidates for this extruder role. Therefore, we selected H-NS(Dorman, 2004), a major NAP, which is capable of promoting the release of DnaA from bound DnaA-boxes in vitro (Fig. S4), to experimentally assess the validity of the extrusion model. To this end, we constructed a hns-titratable strain, incorporating a reporting plasmid to capture changes in DnaA activity across various H-NS expression levels (Fig. 7A). Under steady-state cultivation, increasing aTc concentrations from 0 to 50 ng·ml-1 resulted in hns mRNA levels ranging from 0.25 to 5 times the wild-type levels, without affecting growth rate (Fig. S5 A-B). Within this range, DnaA activity increased by approximately 60 % (Fig. 7B), while the initiation mass decreased by 30 % (Fig. 7C; calculated with Fig. 5S C-D), in qualitative agreement with the predictions of the extrusion model (Fig. S5 E-F).

Titration of hns expression modulates DnaA activity and replication initiation.

(A) Genetic circuit of the hns-titratable strain. Expression of hns is controlled by a Ptet-tetR negative feedback loop integrated at the attB site, with the native hns coding sequence replaced by a kanamycin resistance gene. The plasmid contains Psyn66-mcherry and Pcon-gfp expression cassettes was used to assess DnaA activity. DnaA activity (B) and initiation mass (C) were characterized in M6 medium with varying hns expression levels during steady-state cultivation. (D) mRNA levels of hns (green circles) and dnaA (blue squares) relative to wild-type levels, along with DnaA activity (orange rhombus), were measured during hns shift-up. hns shift-up was induced by the addition of 50 ng·ml−1 aTc at time 0 (dashed line), following steady-state cultivation in M6 medium without aTc. (E) Dynamics of population-averaged oriC number (blue rhombus, left axis) and cellular mass (gray rhombus, right axis) during hns shift-up. Data represent mean ± SD from 3 biological replicates (B-E). (F) Cell cycle-dependent DnaA activity oscillations in hns-titratable cells cultivated in M1 and M6 media under varying aTc concentrations. Relative hns mRNA levels are indicated for each condition. More than 8,000 cells were analyzed for each condition and error bars show the mean ± SEM. (G) Correlation between the volume at maximal DnaA activity (V∗) and the volume at replication initiation (Vi) for hns-titratable cells grown in M1 (green down-pointing triangle) and M6 (red up-pointing triangle) media with varying hns expression levels. Data for dnaA-titratable cells (gray circles) and wild-type dnaA-autoregulated cells (Fig. 5C) are included for comparison.

Since H-NS is a global transcriptional regulator (Hommais et al., 2001), we wonder whether its effect on DnaA activity was mediated through dnaA transcription. To test this, we measured DnaA activity and dnaA mRNA levels during an H-NS shift-up. Upon addition of 50 ng·ml-1 aTc, hns levels increased more than 20-fold within a short period, followed by approximately 80 % increase in DnaA activity, with no significant change in dnaA mRNA levels (Fig. 7D). Additionally, cellular oriC content increased significantly within 30 minutes, while cell mass remained unchanged (Fig. 7E). These results suggest that the rapid increase in DnaA activity and accelerated replication initiation induced by H-NS upregulation is not driven by changes in dnaA transcription.

Leveraging the hns-titratable strain, we re-examined the tight correlation between DnaA activity oscillations and replication initiation. In two different growth media with varying hns levels, the peak of DnaA activity fluctuations (V∗) shifted in concert with the independently measured volume at replication initiation (Vi) (Fig. 7F). Plotting V∗ against Vi, revealed the same equivalency observed in wild-type and dnaA-titratable cells (Fig. 7G). Collectively, these results support the notion that DnaA activity oscillations, modulated by extrusion, coordinate DNA replication and biomass growth.

Discussion

In this study, we proposed chromosome-driven DnaA activity oscillations as a post-translational “rheostat” coordinating DNA replication with biomass growth in E. coli. By developing a SeqA-independent synthetic promoter reporter (Psyn66) on a plasmid, we circumvented confounding factors inherent to native systems (Pountain et al., 2024), thereby achieving the direct quantification of DnaA activity dynamics. Crucially, these oscillations persist when dnaA transcription is fixed, operate independently of SeqA-mediated repression, and peak precisely at replication initiation across all tested growth conditions. This reveals an essential paradigm: bacteria bypass transcriptional control to synchronize genome duplication with growth through regulation of DnaA-chromosome interactions.

Very recently, Iuliani et al. also reported DnaA activity oscillations (Iuliani et al., 2024), which is a significant milestone, innovatively using a chromosomal promoter reporter system and microfluidics for single-cell DnaA activity observations. They revealed key relationships between DnaA activity, cell size, and division. Our work, however, differs in multiple ways. Methodologically, we engineered a dnaA-titratable strain and synthetic promoters like Psyn66, enabling precise dnaA expression control and cleaner DnaA activity reporting. Our mRNA FISH approach uniquely compares DnaA activity and transcription at the single-cell level. Replication initiation volume scales proportionally with peak DnaA activity volume with a slope of 1.0 (R2=0.98, Fig. 7G), indicating predictive correspondence rather than absolute coincidence. While population-level Vi estimation cannot resolve single-cell stochasticity, the consistent V∗: Vi relationship across 20 conditions suggests DnaA activity thresholds predict initiation timing within physiological error margins. Regarding research questions, while Iuliani et al. focused on DnaA-cell size-division relationships, we addressed the decoupling of replication initiation from dnaA expression with our novel extrusion model.

Recently, an integrated “titration-switch” model has been proposed to improve replication cycle stability across different growth rates (Berger & Wolde, 2022; Fu et al., 2023). Although this integrated model predicts a relationship between initiation mass and dnaA mRNA levels that aligns closely with steady-state experimental data (Fig. S6 A), it fails to account for the additional replication initiation observed after dnaA shutdown (Fig. S6 B). By incorporating an extruder into this model, we successfully reproduced the additional replication initiation (Fig. S6 C), thus validating the extrusion concept and aligning with experimental observations. These findings highlight the critical role of extrusion in refining our mechanistic understanding of biomass-DNA coordination.

Mechanistically, we propose that NAPs sense biomass-DNA imbalances and extrude DnaA from chromosomal sites, converting physical displacement into biochemical activation, a spatial control strategy mirroring eukaryotic chromatin regulation. H-NS emerges as a prime extruder candidate given its DNA-condensing ability (Dame et al., 2000; Zimmerman, 2006) and replication-timing effects (Fig. 7). This functional multiplicity likely enhances robustness, allowing cells to maintain replication homeostasis despite fluctuating environments, an evolutionary advantage paralleling eukaryotic checkpoint system.

These findings refine bacterial cell cycle control as a dynamic interplay between physical chromosome remodeling and metabolic sensing. By tethering DnaA activation to biomass accumulation through extrusion, cells achieve an important capability: noise suppression via spatial filtering of transcriptional fluctuations. Such mechanistic parsimony, using chromosome structure to both store genetic information and regulate replication timing, may represent a potential organizational strategy. Indeed, parallels to eukaryotic systems are striking: DnaA extrusion resembles ORC licensing through chromatin accessibility (Fragkos et al., 2015), while biomass-DNA coupling evokes yeast size-control checkpoints (Xie et al., 2022).

While H-NS perturbation supports extrusion mechanism, future work should identify the full extruder interactome and elucidate how metabolic signals modulate their activity. Our synthetic reporter system provides a valuable tool for dissecting the spatial regulation of DNA-binding proteins, with translational potential in antimicrobial development (targeting pathogen-specific extruders) and synthetic biology (engineering tunable replication circuits). More broadly, this work establishes physical chromosome remodeling as a central coordinator of cellular homeostasis, a concept bridging prokaryotic and eukaryotic cell cycle paradigms.

Materials and Methods

Plasmid and strain Construction

Strains used in this study were derived from E. coli MG1655. Information for all strains and plasmids are listed in Tables S1 and S2.

Plasmids for reporting DnaA activity

Plasmids expressing GFP, mCherry, or LacZ under synthetic promoters were constructed based on the pPT plasmid backbone (Zong et al., 2017). The lacZ amplified from the MG1655 genome and mcherry gene amplified from pMD19-hupA-mcherry plasmid, were assembled into the BspQI/BsrGI-digested pPT backbone via the One Step Cloning Kit (Vazyme, C112), to generate the pPT-lacZ and pPT-RFP plasmids, respectively. Synthetic promoters, generated by annealing the primer pairs listed in Table S3, were ligated directly into the BsaI-digested pPT, pPT-RFP or pPT-lacZ plasmid as required. The native dnaA promoter (Pnative) amplified from the MG1655 genome was digested with BsaI, and ligated into the BsaI-digested pPT plasmid. The Pcon-gfp cassette amplified from Pcon-GFP plasmid was assembled into the SpeI-digested Psyn66-RFP plasmid to generate the Psny66-Pcon-FPs plasmid. Primer information for amplifying DNA fragments is detailed in Table S4.

Plasmids for strain construction

For constructing the plasmid pMD19-Rhns, pMD19-RseqA and pMD19-RdnaA, the hns, seqA and dnaA gene were amplified from MG1655 genome, digested, and ligated into the same enzyme-digested pMD19-tetR plasmid. For constructing the plasmid p15A-RdnaA, the T3-Ptet-tetR-dnaA cassette, kanr gene and p15A-T1 fragment were amplified from plasmid pMD19-RdnaA, pEcCas and PZA31-Ptet-M2-GFP, respectively, and then assembled via MultiS One Step Cloning Kit (Vazyme, C113). The plasmid P_CRidnaA1 were obtained by ligating the annealed primer pairs to the BsaI-digested CPP00458 plasmid for the dnaA targeting. The chloramphenicol resistance gene Cmr amplified from the PZA31-Ptet-M2-GFP plasmid replaced the kanr gene in the pEcCas plasmid via λRed recombination method, resulting in the CmPcas plasmid. Primer information is detailed in Table S4.

Strain construction

The seqA-, and hns-titratable strains were constructed as previously described (Zheng et al., 2016). Briefly, after transforming pSIM5 plasmid into MG1655 cells, the Ptet-tetR-driven seqA or hns cassette was inserted at the attB site on the genome, and then the native gene was replaced with the kanr gene amplified from plkml plasmid.

For the construction of dnaA-titratable strain, we first inserted the Ptet-tetR cassette into the intS site via CRISPR/Cas9-λRed genome editing system (Jiang et al., 2015) by targeting 20-nucleotide sequence (tcttcctgcagaccagatcc). Then applying the λRed recombination systems in the CmPcas plasmid, the bla:Ptet-dnaA cassette amplified from the pMD19-RdnaA plasmid via overlap PCR, was inserted between the yidA and yidX genes, and the negative Pnative-dnaA was replaced with the T3-Kan-T1 cassette amplified from the p15A-RdnaA plasmid.

In order to delete the native lacZ gene in the genome of the RdnaA1, Rhns1 and MG1655 cells, the homologous recombination fragment amplified from the genome of CL1 strain, was used to knock out the native lac operon via CRISPR/Cas9-λRed editing system (Jiang et al., 2015) by targeting 20-nucleotide sequence (cttccggctcgtatgttgtg).

For the construction of DnaA shut-down strain (CRidnaA1), the PJ23119-sgRNAdnaA:PJ23100-tetR:Ptet-dUn1Cas12f1 cassette was amplified from the P_CRidnaA1 plasmid and flanked with homologous sequences to the asnA and viaA gene. This cassette was then inserted into the MG1655 genome using CRISPR/Cas9-λRed editing system, targeting the 19-nucleotide sequence (aagccgcctgctcagacgc). The CPP00458 plasmid, kindly provided by the Xiongfei Fu lab, contains the Un1Cas12f1 variant, which carries mutations D143R, T147R, D326A, and K330R that abolish its DNA-cleaving activity. However, it retains the ability to bind DNA in a sgRNA-guided manner, thereby inhibiting the expression of the target gene.

All primers for the construction of DNA fragments are detailed in Table S4.

Growth conditions

Cells for strain construction were grown in Luria-Bertani (LB) medium with appropriate antibiotics: ampicillin (100 µg·ml-1), kanamycin (50 µg·ml-1), spectinomycin (50 µg·ml-1), chloramphenicol (25 µg·ml-1), based on plasmid presents. For titratable strains, 10 ng·ml−1 anhydrotetracycline (aTc) was added during strain construction and seed culture preparation. Seed cultures were initiated by inoculating 3–6 individual colonies from an LB agar plate into 14 ml round-bottom test tubes containing 2 ml of the desired medium for overnight growth.

For physiological phenotypic characterization, all cells were cultivated at 37°C with shaking at 150 rpm in a water bath shaker. The dnaA-, seqA- and hns-titratable cells were characterized under varying aTc concentrations in rich defined medium supplemented with glycerol (M6). hns-titratable cells were also characterized in rich defined medium supplemented with glucose (M1). When strains contained reporter plasmid, 25 µg·ml−1 chloramphenicol was added. The autoregulated-dnaA strain (MGCL1) with a reporter plasmid, was tested in seven different media: M1 (MOPS + EZ + AUCG + Glucose), M2 (MOPS + EZ + Glucose), M3 (MOPS + EZ + AUCG + Gluconate), M6 (MOPS + EZ + AUCG + Glycerol), M12 (MOPS + EZ + Glycerol), M13 (MOPS + CAA + Glycerol), and M18 (MOPS + Glucose). Detailed medium compositions were provided previously (Zheng et al., 2020).

Steady-state growth was established as previously described(Zheng et al., 2020) by culturing cells for ∼10 generations in exponential growth phase with ∼15-fold dilution at OD600 ∼0.2, using the same medium supplemented with antibiotic or aTc as required. OD600 measurements were taken at consecutive time points using a spectrometer (Genesys 10s, Thermo Fisher Scientific), and the growth rate (or doubling time) λ was calculated by fitting the data to an exponential growth curve. Samples for determination of population-averaged cell mass and cellular origin number (ō), quantification of mRNA levels, and mRNA fluorescence in situ hybridization (FISH) were collected when OD600 reached ∼0.2.

Measurements of population-averaged cellular mass , cellular oriC number , the initiation to division period (C+D), and initiation mass (mi)

The population-averaged cell mass , cellular oriC number (ō), and initiation mass (mi) were calculated as described previously(Zheng et al., 2020). Briefly, was determined by dividing the OD600 by the cell concentration (OD600 ml per 109 cells). For cell counting, samples were collected in precooled cell count buffer (0.9% NaCl with 0.12% formaldehyde, filtered with a 0.22 µm filter) and stained with 1 µg·ml−1 DAPI in the same buffer. Cell number concentration was measured by counting the DAPI-stained particles using flow cytometry equipped with 405 nm laser (CytoFLEX, Beckman Coulter Life Sciences). The ō was calculated based on the classical run-out experiments(Zheng et al., 2020). In brief, cell suspensions were treated with 500 µg·ml−1 rifampicin and 30 µg·ml−1 cephalexin to inhibit new initiation and cell division, respectively, and maintained for 2-3 mass doublings to complete ongoing replication. Cells were fixed with 70% precooled absolute ethanol, and treated with stain buffer (20mM Tris-HCl, pH 8.0, 130 mM NaCl, 10 ng·ml−1 DAPI). Using the same flow cytometry, we measured the cellular origin distribution based on DAPI signaling, from which ō was calculated. C+D was calculated based on the equation: C+D = ln(ō)/λ. mi was calculated based on the equation: (Bremer et al., 1979; Si et al., 2017; Zheng et al., 2016).

Screening of the synthetic promoter library

After transforming the reporter plasmid into dnaA-titratable cells, the seed culture was cultivated in M6 medium with 10 ng·ml-1 aTc, then diluted 1:180 in fresh medium containing either 0.5 or 50 ng·ml-1 aTc and cultivated at 37 °C and 800 r.p.m. in a digital thermo-shaker (AOSHENG) using flat-bottom 96-well plates (Corning). Following approximately 4 hours of pre-cultivation, the cultures were diluted 1:180 in fresh medium with the same aTc concentration. The OD600 and GFP fluorescence intensity were continuously monitored during cell growth using a Synergy H1 microplate reader (BioTek). The non-fluorescent strain (MG1655) cultured in 180 ul of M6 medium for each experiment was used to subtract the background fluorescence. For OD600 between 0.1 and 0.2, the repression fold change was calculated by dividing the mean fluorescence intensity per OD600 of the cells grown with 0.5 ng·ml-1 aTc induction by the corresponding intensity for cells grown with 50 ng·ml-1 aTc induction.

Quantitative real-time PCR

Quantitative real-time PCR was performed using a protocol similar to previously described (Zheng et al., 2016). Briefly, RNA was stabilized using Bacteria Protect Reagent (Qiagen) and extracted using RNeasy mini kit (Qiagen). 500 ng RNA was reverse transcribed (TAKARA), and the resulting cDNA samples were diluted 1:25 with PCR-grade water. Quantitative real-time PCR was carried out on a qTower3 (Analytikjena) or CFX Connect system (Bio-Rad), using the following program: 30 s at 95 °C, followed by 40 cycles of denaturation (5 s at 95 °C), annealing, and elongation (30 s at 58 °C). Primer sequences were listed in Table S4. The rpoA gene was used as the reference gene.

lacZ and dnaA mRNA fluorescence in situ hybridization (FISH)

FISH for lacZ or dnaA mRNA was performed using a method similar to previously described (Skinner et al., 2013). Briefly, antisense DNA oligo probes were designed against lacZ or dnaA, and synthesized with TAMRA labeling at the 3’ end. The sequences of probes targeting lacZ or dnaA are either as previously described(Skinner et al., 2013) or listed in Table S5. 27 ml steady-state growth culture (OD600∼0.2) was fixed by direct adding 3 ml of 37 % (vol/vol) formaldehyde, following by incubation for 30 min at room temperature with gentle mixing using a nutator. The fixed cells were collected by centrifugation at 600 g for 8 minutes, and washed twice with 1 ml RNase-free PBS. The cell pellet was resuspended in 300 µl DEPC water, and 700 µl precooled absolute ethanol was added to permeabilize the cells for 1 hour at room temperature in the nutator. After permeabilization, the cells were divided into two equal parts for characterization of lacZ or dnaA mRNA as needed. Hybridization was performed as described previously(Skinner et al., 2013), with the stringency of the hybridization and wash solutions adjusted by changing the concentration of formamide, 30 % for dnaA probes and 40 % for lacZ probes. After hybridization, the cells were washed with the appropriate wash solution and prepared for imaging.

Cell imaging and cell size parameters measurement

For imaging, 3 µl of cell suspension was pipetted onto the center of a 24×60-mm coverslip. A thin 1.5 % agarose pad (prepared with 1xPBS) was then placed on top of the cell suspension droplet. Fluorescent and phase images were acquired using an inverted microscope (IX-83, Olympus) equipped with a 100x oil objective (Olympus), an automated xy-stage (ASI, MS2000), and a sCMOS camera (Prime BSI, photometrics). Fluorescence excitation was performed as previously described (Xia et al., 2022). Solid-state lasers (Coherent OBIS: 405–100 LX, 561–100 LS) and Semrock emission bandpass filters (445/40 nm for DAPI and 598/25 nm for RFP) were used for excitation and collecting fluorescent signals.

Images were aligned with phase contrast channel by selecting at least six pairs of common control points and applying a geometric transformation using the MATLAB imwarp function. This allowed for adjustment of the cell masks in the RFP or DAPI channel. A customized ImageJ -based image-processing package, MicrobeJ (v5.13l1) plugin for Fiji (Ducret et al., 2016), was used to contour cells and calculate cell size parameters and fluorescence intensity, including the mean cell width, cell length, and cell area from phase-contrast images. The cell volume was calculated using a cylinder-plus-two-hemispheres geometry. The total RFP intensity per cell volume, corrected for background subtraction, was then used to estimate the mRNA concentration of lacZ or danA.

Characterization of cell cycle-dependent fluctuations in DnaA activity

Data filtering was first performed based on the intensity of the DAPI fluorescence signal. To eliminate non-cell regions resulting from incorrect segmentation of the phase contrast images, only DAPI-positive cells were included in the analysis. To further minimize the potential impact of variability in fixation, permeabilization, and wash efficiencies, which could affect the TAMRA fluorescence signal (representing the mRNA FISH signal), only cells with mean DAPI signals within the 95% confidence intervals were included in the subsequent analysis.

The concentration of the lacZ mRNA ([mZ]) can be expressed as [mZ] = J/δZ, where δZ is the degradation rate and J is the mRNA synthesis flux. J is in turn given by the promoter on-rate (k), the RNAP concentration ([RNAP]), and the concentration of the lacZ gene ([GZ]). Assuming that the factors δZ, [RNAP], and [GZ] are the same between the two strains with Psyn66 and Pcon driving lacZ, then the ratio of the mRNA concentration obtained at each point of the cell cycle gives the ratio of the promoter on-rates, kPsyn66 and kPcon. Since a constitutive promoter is non-responsive to DnaA activity, kcon is independent of cell cycle and we can obtain DnaA activity regulated promoter activity as , then the DnaA activity is denoted as .

The concentration of the lacZ mRNA were binned based on cell volume, with a bin size of 0.1 µm3, was used to determine [mZ](Psyn66) and [mZ](Pcon). The relation between [mZ](Pcon) and cell volume was smoothed using the MATLAB smooth function, and from this, ksyn66 was deduced.

Based on the DnaA activity oscillations, the cell volume where DnaA activity reaches maximum (V∗) was calculated during a representative birth-to-division cell cycle. The cell volume distribution was used to calculate the number of cells in each volume-doubling range. The range containing the highest cells was considered the representative birth-to-division cell cycle and was shaded in gray (Fig. S2 D). From this, V∗ was obtained.

Calculation of cell volume at the time of replication initiation (Vi)

Cell volume at the time of replication initiation (Vi) was calculated based on population-averaged cell volume , and the population-averaged cellular oriC number (ō). The equation used for this calculation is as follows: where vi is the initiation volume by definition, i.e. cell volume per oriC at the time of replication initiation(Bremer et al., 1979; Si et al., 2017; Zheng et al., 2016). was obtained by analyzing the phase contrast images, ō was obtained by run-out experiments (Zheng et al., 2020). is the cellular oriC number at the time of replication initiation. ⌊log2 ō⌋ denotes the floor function of log2 ō.

DnaA shut-down experiment

To shut-down DnaA expression, steady-state growth of the CRidnaA1 cells was first established after ∼10 generations of exponential growth in M1 medium. At OD600 between 0.15 and 0.2, 24 ml of cell culture was transferred to 12 ml of pre-warmed M1 medium containing 150 ng·ml−1 aTc to induce dUn1Cas12f1 expression at time 0. Under the guidance of sgRNAdnaA, the transcription of dnaA gene was inhibited. Samples were collected at various time points to determine population-averaged cellular mass , cellular oriC number (ō), and to quantify mRNA levels. When the OD600 reached around 0.2, the cultures were diluted about 2-fold with pre-warmed M1 medium containing 50 ng·ml−1 aTc to maintain adequate nutrition. The steady-state growing MG1655 cells in M1 medium were also collected for quantification of wild-type dnaA mRNA levels. Three independent experiments were conducted.

H-NS shift-up experiment

The DnaA activity reporter plasmid Psyn66-Pcon-FPs was transformed into the hns-titratable strain (Rhns2), and cells were cultivated overnight in M6 medium containing 10 ng·ml−1 aTc and 25 µg·ml−1 chloramphenicol for seed culture. Seed cultures were washed and grown in the same medium without aTc for ∼10 generations of exponential growth. At an OD600 of ∼0.2, 40 ml of cell culture was transferred to 20 ml of pre-warmed same medium containing 150 ng·ml−1 aTc to induce H-NS expression at time 0. Samples were collected at time points -2, 1, 3, 5, 7, 10, 15, 20, 25, and 30 minutes for the determination of population-averaged cellular mass , cellular oriC number (ō), and quantification of mRNA levels. Steady-state growing MG1655 in M6 medium were also collected for hns mRNA quantification. Three independent experiments were conducted.

Electrophoretic mobility shift assay (EMSA)

The genes encoding dnaA and hns from the MG1655 genome were individually amplified and cloned into the expression vector pET-28a using primers listed in Table S4. E. coli BL21 (DE3) carrying pET-28a-DnaA or pET-28a-H-NS was used for overproduction of His6-DnaA or His6-H-NS. For each protein purification, a 1 L culture was induced by the addition of 1 mM IPTG at OD600 = 0.4. After 1.5 hours of incubation, cells were harvested by centrifugation (5,000 rpm for 20 min at 4 °C). The pellet was resuspended in 50 ml of buffer A (50 mM Tris-HCl, pH 7.0, 500 mM NaCl, 0.5 mM PMSF, 5 % glycerol). Cell resuspension was lysed on ice for 30 minutes (3 seconds on, 7 seconds off) using a sonicator at 4 °C, and cell debris was removed by centrifugation (14,000 rpm for 60 min at 4 °C). Subsequently, the protein was purified from the soluble supernatant by Ni2+-affinity chromatography (HisTrap™ HP Columns; GE Healthcares) with a ÄKTA Pure Protein Purification System. Proteins were collected using gradient elution with buffer B [20 mM Tris-HCl, pH 7.5, 500 mM NaCl, 400 mM imidazole, 5 % glycerol]. Eluted proteins were further purified and concentrated by ultrafiltration using Amicon Ultra 10K (for H-NS) and 30K (for DnaA) filters (Millipore). Finally, after overnight dialysis (0.025 µm Membrane Filter; MF-Millipore) at 4 °C, the protein was stored in buffer C [45 mM Hepes-KOH, pH 7.6, 600 mM potassium glutamate, 1 mM DTT, 10 mM magnesium acetate, 0.5 mM EDTA, 20 % glycerol]. Judgement by SDS/PAGE with Coomassie Brilliant blue staining, the purity of each protein was >90 %. For the experiments where H-NS acts on the datA bound by DnaA, the purified DnaA, datA and ATP were first incubated in buffer D (20 mM Tris-HCl, pH 7.5, 150 mM potassium glutamate, 10 mM magnesium acetate, 8 mM DTT) at 20 °C for 10 min. Then the H-NS was added to buffer D (15 µl) which contained DnaA-datA complexes and incubated for additional 10 min at 20 °C. The final complexes were separated by electrophoresis in 2 % (wt/vol) agarose gel at 80 V for 70 min in Tris-Borate buffer, followed by staining with SYBR Safe (Invitrogen).

Titration model

In the titration model, we posit that the cell mass grows exponentially according to the differential equation , where V represents the volume of the cell and λ = 1.54 h−1(corresponding to doubling time of 27 min) is the growth rate. Concurrently, the number of total DnaA protein (A) increases at a rate proportional to the cell’s mass, described by , with α A= 300 h−1µm−3 denoting the synthesis rate per unit volume. The chromosome’s structure is abstracted as a vector representing the copy numbers of uniformly distributed sites along its length, and the number of DnaA-boxes (Ab) is quantified by summing the boxes number at their respective locations. In this model, 300 DnaA-boxes are evenly distributed on the chromosome while an additional 100 DnaA-boxes are located near the oriC. This arrangement approximates a head-weighted distribution of DnaA-boxes and helps prevent over-initiation(Hansen et al., 1991). Binding of DnaA proteins to these DnaA-boxes allows us to calculate the concentration of free DnaA as [Af] = (A − Ab)/V, under condition of (A − Ab) ≥ 0. DNA replication commences once the concentration of free DnaA exceeds a critical threshold . 10 min sequestration time was set after replication initiation to mimic the sequestration time(Campbell & Kleckner, 1990).

To explore the dynamics of the oriC number following the cessation of dnaA expression, we can set the synthesis rate to be zero αA = 0. To simulate the dependence of initiation mass on dnaA expression, αA varies from 60 h−1µm−3 to 3840 h−1µm−3.

Extrusion model

The extrusion model was formulated following the titration model, with an additional extruder introduced as a protein with synthesis rate αH = 180 h−1µm−3 and dynamics dH/dt = αHV. The additional extruder competes the DnaA-boxes with DnaA proteins, with higher affinity to DnaA-boxes. So that it can displace DnaA from the DnaA-boxes, leading to a revised calculation for free DnaA concentration as [Af] = (A − (Ab − H))/V, under condition of (Ab − H) ≥ 0 and (A − (Ab − H)) ≥ 0. To get stochastic modelling for adder phenomenon, we introduced white noises on cell mass growth rate and the DnaA synthesis rate at each replication cycle: λ = λ0 + ηλ0, αA = α0 + ηα0 where λ0 = 0.66 h−1, α0 = 300 h−1µm−3 and η = 0.1 representing white noise. This extrusion model exhibits consistent performance under various parameter ranges.

DnaA-ATP/DnaA-ADP switch model

The switch model was simulated following the idea of LDDR model proposed by Berger et al. (Berger & Wolde, 2022). The formulations of the DnaA-ATP fraction dynamics is: In this equation [D]T = 400µm−3 is the concentration of total DnaA proteins in a cell αl = 750h−1 is the activation rate of DnaA-ATP by lipid; αd1 = 100 h−1, αd2 = 643 h−1 are the activation rate of DnaA-ATP by the DARS1 and DARS2 sites; βdatA = 600 h−1, βrida = 500 h−1 are the deactivation rate of DnaA-ATP by datA and RIDA; KD = 50µm−3 the dissociation constant of DnaA activation and deactivation; Nd1, Nd2, No are copy number of DARS1, DARS2 and oriC sites extracted dynamically from the DNA copy number vector; αA = 300 h−1µm−3 is the DnaA expression level. The DNA replication initiation was determined when f reaches a threshold fC = 0.75 (Berger & Wolde, 2022). To simulate the dnaA shut-down, we simply deleted the term of dnaA synthesis (λ(1 − f)).

Titration-switch model and titration-switch-extrusion model

In the integrated model combining titration and switch mechanisms, replication initiation is triggered when the concentration of free DnaA-ATP exceeds a threshold . The concentration of free DnaA-ATP is titrated by the DnaA-boxes on the chromosome as in the titration model, with , under condition of (AATP − Ab) ≥ 0. The parameters related to the titration dynamics remain consistent with those described in earlier sections of titration model.

The dynamics of total DnaA-ATP (AATP) is governed by both expression and conversion processes as described in Berger et al. (Berger & Wolde, 2022). According to this framework, the lipid and DARS systems facilitate the conversion of DnaA-ADP to DnaA-ATP at constant rates, while the datA site and RIDA systems also play a role in converting DnaA-ATP to DnaA-ADP, again at fixed rates. Additionally, the synthesis of new DnaA proteins provides another source of DnaA-ATP.

To be specific, DNA replication initiates was set when the concentration of free DnaA-ATP, exceeds a threshold . The dynamics of total DnaA number in a cell A followed , and the dynamics of total DnaA-ATP number AATP followed: where the three terms on the right represent activation and deactivation of DnaA-ATP and new synthesis of DnaA-ATP. In these equations is the concentration of DnaA-ADP; [DT] = 400 µm−3 is the total DnaA concentration; is the dissociation constant of DnaA promoter; ϕp0 is the DnaA protein ratio relative to all proteins, which is proportional to αA and ranges from 0.002% to 0.1%; n = 5 is the cooperativity of dnaA expression; The same activation/deactivation rates and dissociation constants of DnaA (αl, αd1, αd2, βdatA, βrida, KD) are used as in the switch model; Nd1, Nd2, No are copy number of DARS1, DARS2 and oriC sites extracted dynamically from the DNA copy number vector. To get a good fit of the initiation mass-dnaA expression relation, high threshold of free DnaA-ATP concentration was applied,.

To simulate the titration-switch-extrusion model, we incorporated the extruder as described in the extrusion model, where its amount was given by H = αHV. The concentration of free DnaA-ATP was then calculated as , under condition of (Ab − H) ≥ 0 and (AATP − (Ab − H)) ≥ 0. To explore the best fit of continuous DNA replication rounds following dnaA shut-down, αH = 550 h−1µm−3 is used.

All parameters are summarized in Tables S6.